\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{tcolorbox}
\usepackage{tikz}
\geometry{a4paper, margin=1in}

\usetikzlibrary{arrows.meta, positioning, calc, decorations.pathreplacing}

\title{Arithmetic Coding and Range Coding: A Detailed Tutorial}
\author{}
\date{}

\begin{document}

\maketitle

\section{Arithmetic Coding and Range Coding: A Detailed Tutorial}

\subsection{Introduction}

Entropy coding is the final stage in many compression systems.
Instead of assigning fixed codewords to symbols (like Huffman coding),
\textbf{Arithmetic Coding} and \textbf{Range Coding} represent the entire
message as a single number inside the unit interval $[0,1)$.
This allows code lengths to approach the theoretical entropy limit.

\begin{tcolorbox}[colback=blue!3,colframe=blue!40!black,title=Key Idea]
Arithmetic coding does not assign codes to symbols individually.
Instead, it assigns a code to the entire \textit{sequence} by repeatedly shrinking an interval.
The more probable the message, the larger the final interval and the fewer bits needed.
\end{tcolorbox}

We now explain arithmetic coding in detail, followed by a complete worked
example, a comparison with range coding, and finally the same example solved
using range coding.

\subsection{Arithmetic Coding: Theory and Algorithm}

Arithmetic coding encodes a message by repeatedly narrowing an interval in $[0,1)$.
Each symbol reduces the interval based on its probability.

\begin{tcolorbox}[colback=blue!3,colframe=blue!40!black,title=Algorithm: Arithmetic Coding Encoding]
\textbf{Step 0: Initialization} \\
Set $\text{low} = 0$ and $\text{high} = 1$.

\vspace{2mm}
\textbf{Step 1: Compute cumulative frequencies} \\
Given an alphabet $\{s_1, s_2, \dots, s_k\}$ with probabilities $p(s_i)$, compute:
\[
C(s_1) = 0, \qquad C(s_i) = \sum_{j < i} p(s_j) \text{ for } i = 2, 3, \dots, k
\]

\vspace{2mm}
\textbf{Step 2: For each symbol in the message to be encoded} \\
For each symbol $s$ (taken in order from the message), perform:
\begin{align*}
\text{range} &= \text{high} - \text{low} \\
\text{high} &= \text{low} + \text{range} \times (C(s) + p(s)) \\
\text{low} &= \text{low} + \text{range} \times C(s)
\end{align*}

\vspace{2mm}
\textbf{Step 3: Generate the final code} \\
After encoding all symbols, any number $x$ in the interval $[\text{low}, \text{high})$ represents the entire message. Typically, we output:
\[
\text{code} = \frac{\text{low} + \text{high}}{2} \quad \text{(the midpoint of the final interval)}
\]
\end{tcolorbox}

\subsection{Example: Arithmetic Coding Step-by-Step}

Consider the alphabet $\{A,B,C\}$ with frequencies:

\[
f(A)=5,\qquad f(B)=3,\qquad f(C)=2,\qquad \text{Total}=10.
\]

Thus probabilities:

\[
p(A)=0.5,\qquad p(B)=0.3,\qquad p(C)=0.2.
\]

We want to encode the message:

\[
\text{MESSAGE} = \texttt{BAC}
\]

\subsubsection*{Step 0: Initialization}
\[
\text{low} = 0,\qquad \text{high} = 1
\]

\subsubsection*{Step 1: Compute cumulative frequencies}
The cumulative frequencies define the sub-interval boundaries and are sorted in alphabetical order:

\[
C(A) = 0,\qquad C(B) = p(A) = 0.5,\qquad C(C) = p(A) + p(B) = 0.8
\]

\subsubsection*{Step 2: Encode each symbol in the message}

\textbf{Symbol 1: B}

\begin{align*}
\text{range} &= \text{high} - \text{low} = 1 - 0 = 1 \\
\text{high} &= \text{low} + \text{range} \times (C(B) + p(B)) = 0 + 1 \times (0.5 + 0.3) = 0.8 \\
\text{low} &= \text{low} + \text{range} \times C(B) = 0 + 1 \times 0.5 = 0.5
\end{align*}

Interval after encoding B: $[0.5,\; 0.8)$

\vspace{2mm}
\textbf{Symbol 2: A}

\begin{align*}
\text{range} &= \text{high} - \text{low} = 0.8 - 0.5 = 0.3 \\
\text{high} &= \text{low} + \text{range} \times (C(A) + p(A)) = 0.5 + 0.3 \times (0 + 0.5) = 0.5 + 0.15 = 0.65 \\
\text{low} &= \text{low} + \text{range} \times C(A) = 0.5 + 0.3 \times 0 = 0.5
\end{align*}

Interval after encoding A: $[0.5,\; 0.65)$

\vspace{2mm}
\textbf{Symbol 3: C}

\begin{align*}
\text{range} &= \text{high} - \text{low} = 0.65 - 0.5 = 0.15 \\
\text{high} &= \text{low} + \text{range} \times (C(C) + p(C)) = 0.5 + 0.15 \times (0.8 + 0.2) = 0.5 + 0.15 \times 1.0 = 0.65 \\
\text{low} &= \text{low} + \text{range} \times C(C) = 0.5 + 0.15 \times 0.8 = 0.5 + 0.12 = 0.62
\end{align*}

Final interval after encoding all symbols: $[0.62,\; 0.65)$

\subsubsection*{Step 3: Generate the final code}

Any number in $[0.62, 0.65)$ encodes the message \texttt{BAC}. We choose the midpoint:

\[
\text{code} = \frac{0.62 + 0.65}{2} = 0.635
\]

This is the arithmetic-coded representation of \texttt{BAC}. In practice, we would output this number in binary form.

\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=1.3, 
    every node/.style={font=\small},
    interval/.style={thick},
    selected/.style={very thick, blue!70!black},
    annotation/.style={font=\footnotesize, text=gray}
]

% Initial interval
\node[anchor=east] at (-0.8,0) {\textbf{Step 0}};
\draw[interval] (0,0) -- (10,0);
\draw[annotation] (0,-0.15) -- (0,0);
\draw[annotation] (10,-0.15) -- (10,0);
\node[below] at (0,-0.1) {$0$};
\node[below] at (10,-0.1) {$1$};
\node[above] at (5,0.2) {Initial Interval $[0,1)$};

% Step 1: Encode B → [0.5, 0.8)
\node[anchor=east] at (-0.8,-1.4) {\textbf{Step 1}};
\node[annotation, right] at (-0.8,-1.4) {Encode B};
\draw[interval] (0,-1.4) -- (10,-1.4);
\draw[selected] (5,-1.4) -- (8,-1.4);
\draw[annotation] (5,-1.55) -- (5,-1.4);
\draw[annotation] (8,-1.55) -- (8,-1.4);
\node[below] at (5,-1.5) {$0.5$};
\node[below] at (8,-1.5) {$0.8$};
\node[above, blue!70!black] at (6.5,-1.2) {$[0.5,\;0.8)$};

% Step 2: Encode A → [0.5, 0.65)
\node[anchor=east] at (-0.8,-2.8) {\textbf{Step 2}};
\node[annotation, right] at (-0.8,-2.8) {Encode A};
\draw[interval] (0,-2.8) -- (10,-2.8);
\draw[selected] (5,-2.8) -- (6.5,-2.8);
\draw[annotation] (5,-2.95) -- (5,-2.8);
\draw[annotation] (6.5,-2.95) -- (6.5,-2.8);
\node[below] at (5,-2.9) {$0.5$};
\node[below] at (6.5,-2.9) {$0.65$};
\node[above, blue!70!black] at (5.75,-2.6) {$[0.5,\;0.65)$};

% Step 3: Encode C → [0.62, 0.65)
\node[anchor=east] at (-0.8,-4.2) {\textbf{Step 3}};
\node[annotation, right] at (-0.8,-4.2) {Encode C};
\draw[interval] (0,-4.2) -- (10,-4.2);
\draw[selected] (6.2,-4.2) -- (6.5,-4.2);
\draw[annotation] (6.2,-4.35) -- (6.2,-4.2);
\draw[annotation] (6.5,-4.35) -- (6.5,-4.2);
\node[below] at (6.2,-4.3) {$0.62$};
\node[below] at (6.5,-4.3) {$0.65$};
\node[above, blue!70!black] at (6.35,-4.0) {$[0.62,\;0.65)$};
\node[above, red!70!black] at (6.35,-3.8) {Final code $=0.635$};

% Add a visual guide showing the shrinking process
\draw[decorate, decoration={brace, amplitude=4pt, mirror}, gray!50] 
    (10.2,0) -- (10.2,-4.2) node[midway, right, xshift=3pt] 
    {Interval shrinks with each symbol};

% Add probability annotation
\node[annotation, text=gray!60, align=left] at (11.5,-2.1) 
    {Symbol probabilities:\\ 
    $p(A)=0.5$, $p(B)=0.3$, $p(C)=0.2$};

\end{tikzpicture}
\caption{Arithmetic Coding Interval Refinement for Message \texttt{BAC}. 
Each step selects a sub-interval based on the symbol being encoded. 
The final code $0.635$ is the midpoint of the final interval $[0.62,0.65)$.}
\label{fig:arithmetic}
\end{figure}

\subsection{Why People Prefer Range Coding}

Arithmetic coding is theoretically elegant but has practical implementation challenges:

\begin{itemize}
    \item \textbf{Arbitrary precision requirement:} As encoding progresses, the interval becomes very small. To represent the final interval precisely, we need high-precision arithmetic. For long messages, this requires arbitrary-precision numbers or specialized libraries.
    
    \item \textbf{Incremental output complexity:} In practice, we don't want to wait until the entire message is encoded before outputting bits. Instead, we output bits incrementally as they become fixed. This process—often called \textit{normalization} or \textit{bit-stuffing}—requires careful handling of carry propagation.
    
    \item \textbf{Computational cost:} Multiplication and division operations at each step are more expensive than the integer operations used in range coding.
    
    \item \textbf{Historical patent issues:} Arithmetic coding was subject to patents for many years, discouraging its use in some applications (though this is no longer an issue today).
\end{itemize}

\begin{tcolorbox}[colback=yellow!5,colframe=orange!70!black,title=Normalization in Arithmetic Coding]
Unlike floating-point numbers that can underflow to zero, the interval in arithmetic coding never underflows—it simply becomes very small. The term \textit{normalization} in arithmetic coding refers to the process of \textbf{outputting bits incrementally} as they become known, not to preventing precision loss. 

For example, after encoding the first few symbols, the interval $[0.5, 0.8)$ tells us that the first bit of the output will be \texttt{1} (since the entire interval lies in the top half of $[0,1)$). We can output that bit immediately and double the interval to continue encoding.

If we are content to store the entire final fraction (like $0.635$) without incremental output, no normalization is needed. However, for long messages, storing the full precision fraction becomes impractical, which is why incremental output is used in real implementations.
\end{tcolorbox}

\textbf{Range Coding} addresses these challenges:
\begin{itemize}
    \item Uses fixed-precision integer arithmetic (e.g., 32-bit or 64-bit integers)
    \item Handles incremental output naturally through renormalization
    \item Uses only integer operations (division, multiplication, shifts)
    \item Faster on most hardware architectures
\end{itemize}

Range coding is simply a \textbf{reparameterization of arithmetic coding} designed for efficient implementation.

\subsection{Range Coding: Theory and Algorithm}

Range coding works on integers rather than real-number intervals. The fundamental insight is that we can scale the $[0,1)$ interval to a large integer range, e.g., $[0, 2^{32})$. All operations then become integer arithmetic.

\begin{tcolorbox}[colback=blue!3,colframe=blue!40!black,title=Algorithm: Range Coding Encoding]
\textbf{Step 0: Initialization} \\
Set $\text{low} = 0$ and $\text{range} = R$ (a large integer, typically $2^{32}$).

\vspace{2mm}
\textbf{Step 1: Compute cumulative frequencies} \\
Given an alphabet $\{s_1, s_2, \dots, s_k\}$ with frequencies $f(s_i)$ and total frequency $\text{total} = \sum_i f(s_i)$, compute:
\[
C(s_1) = 0, \qquad C(s_i) = \sum_{j < i} f(s_j) \text{ for } i = 2, 3, \dots, k
\]

\vspace{2mm}
\textbf{Step 2: For each symbol in the message to be encoded} \\
For each symbol $s$ (taken in order from the message), perform:
\begin{enumerate}
    \item Compute $\text{unit} = \text{range} / \text{total}$ (integer division)
    \item Update $\text{low} = \text{low} + \text{unit} \times C(s)$
    \item Update $\text{range} = \text{unit} \times f(s)$
    \item \textbf{Renormalization:} While $\text{range} < 2^{24}$ (or other threshold), output the top byte of $\text{low}$ and shift $\text{low}$ and $\text{range}$ left by 8 bits.
\end{enumerate}

\vspace{2mm}
\textbf{Step 3: Generate the final code} \\
After encoding all symbols, output the remaining bytes of $\text{low}$ to complete the compressed representation.
\end{tcolorbox}

\subsection{Range Coding Example: Same Message (BAC)}

We now encode the same message \texttt{BAC} using integer range coding. We'll use a smaller initial range of $R = 2^{16} = 65536$ for simplicity in our calculations. The process is identical with $2^{32}$.

\subsubsection*{Step 0: Initialization}
\[
\text{low} = 0,\qquad \text{range} = 65536
\]

\subsubsection*{Step 1: Compute cumulative frequencies}
\[
f(A)=5,\quad f(B)=3,\quad f(C)=2,\quad \text{total}=10
\]
\[
C(A)=0,\quad C(B)=5,\quad C(C)=8
\]

\subsubsection*{Step 2: Encode each symbol in the message}

\textbf{Symbol 1: B}

\begin{align*}
\text{unit} &= \frac{65536}{10} = 6553 \quad \text{(integer division truncates)} \\
\text{low} &= 0 + 6553 \times 5 = 32765 \\
\text{range} &= 6553 \times 3 = 19659
\end{align*}

Renormalization check: $19659 \ge 2^{14}$? No renormalization needed for this simple example.

\textbf{Real-number equivalent:} $\left[\frac{32765}{65536}, \frac{32765+19659}{65536}\right) \approx [0.5000, 0.8000)$

\vspace{2mm}
\textbf{Symbol 2: A}

\begin{align*}
\text{unit} &= \frac{19659}{10} = 1965 \\
\text{low} &= 32765 + 1965 \times 0 = 32765 \\
\text{range} &= 1965 \times 5 = 9825
\end{align*}

Renormalization check: $9825 \ge 2^{14}$? No renormalization needed.

\textbf{Real-number equivalent:} $\left[\frac{32765}{65536}, \frac{32765+9825}{65536}\right) \approx [0.5000, 0.6500)$

\vspace{2mm}
\textbf{Symbol 3: C}

\begin{align*}
\text{unit} &= \frac{9825}{10} = 982 \\
\text{low} &= 32765 + 982 \times 8 = 32765 + 7856 = 40621 \\
\text{range} &= 982 \times 2 = 1964
\end{align*}

Renormalization check: $1964 \ge 2^{14}$? No renormalization needed.

\textbf{Real-number equivalent:} $\left[\frac{40621}{65536}, \frac{40621+1964}{65536}\right) \approx [0.6198, 0.6498)$

\subsubsection*{Step 3: Generate the final code}

The final integer interval is $[\text{low}, \text{low}+\text{range}) = [40621, 42585)$.
Any integer in this range encodes the message. We output the value $40621$ as our compressed representation.

\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=1.3,
    every node/.style={font=\scriptsize},
    interval/.style={thick},
    selected/.style={very thick, blue!70!black},
    annotation/.style={font=\footnotesize, text=gray}
]

% Moderate spacing
\def\stepheight{1.7}

% Initial interval
\node[anchor=east] at (-1.2,0) {\textbf{Step 0}};
\draw[interval] (0,0) -- (10,0);
\draw[annotation] (0,-0.15) -- (0,0);
\draw[annotation] (10,-0.15) -- (10,0);
\node[below] at (0,-0.1) {$0$};
\node[below] at (10,-0.1) {$65536$};
\node[above] at (5,0.2) {Initial Range $[0,65536)$};

% Step 1: Encode B → [32765, 52424)
\node[anchor=east] at (-1.2,-\stepheight) {\textbf{Step 1}};
\node[annotation, right] at (-1.2,-\stepheight) {Encode B};
\draw[interval] (0,-\stepheight) -- (10,-\stepheight);
\draw[selected] (5,-\stepheight) -- (8,-\stepheight);
\draw[annotation] (5,-\stepheight-0.12) -- (5,-\stepheight);
\draw[annotation] (8,-\stepheight-0.12) -- (8,-\stepheight);
\node[below] at (5,-\stepheight-0.08) {$32765$};
\node[below] at (8,-\stepheight-0.08) {$52424$};
\node[above, blue!70!black] at (6.5,-\stepheight+0.2) {$[32765,\;52424)$};

% Real-number equivalent placed to the RIGHT
\node[annotation, text=teal!60!black, anchor=west] at (8.5,-\stepheight) 
    {$\approx [0.5000,\;0.8000)$};

% Step 2: Encode A → [32765, 42590)
\node[anchor=east] at (-1.2,-2*\stepheight) {\textbf{Step 2}};
\node[annotation, right] at (-1.2,-2*\stepheight) {Encode A};
\draw[interval] (0,-2*\stepheight) -- (10,-2*\stepheight);
\draw[selected] (5,-2*\stepheight) -- (6.5,-2*\stepheight);
\draw[annotation] (5,-2*\stepheight-0.12) -- (5,-2*\stepheight);
\draw[annotation] (6.5,-2*\stepheight-0.12) -- (6.5,-2*\stepheight);
\node[below] at (5,-2*\stepheight-0.08) {$32765$};
\node[below] at (6.5,-2*\stepheight-0.08) {$42590$};
\node[above, blue!70!black] at (5.75,-2*\stepheight+0.2) {$[32765,\;42590)$};

% Real-number equivalent placed to the RIGHT
\node[annotation, text=teal!60!black, anchor=west] at (8.5,-2*\stepheight) 
    {$\approx [0.5000,\;0.6500)$};

% Step 3: Encode C → [40621, 42585)
\node[anchor=east] at (-1.2,-3*\stepheight) {\textbf{Step 3}};
\node[annotation, right] at (-1.2,-3*\stepheight) {Encode C};
\draw[interval] (0,-3*\stepheight) -- (10,-3*\stepheight);
\draw[selected] (6.2,-3*\stepheight) -- (6.5,-3*\stepheight);
\draw[annotation] (6.2,-3*\stepheight-0.12) -- (6.2,-3*\stepheight);
\draw[annotation] (6.5,-3*\stepheight-0.12) -- (6.5,-3*\stepheight);
\node[below] at (6.0,-3*\stepheight-0.08) {$40621$};
\node[below] at (6.7,-3*\stepheight-0.08) {$42585$};
\node[above, blue!70!black] at (6.35,-3*\stepheight+0.2) {$[40621,\;42585)$};
\node[above, red!70!black] at (6.35,-3*\stepheight+0.45) {Final code $=40621$};

% Real-number equivalent placed to the RIGHT
\node[annotation, text=teal!60!black, anchor=west] at (8.5,-3*\stepheight) 
    {$\approx [0.6198,\;0.6498)$};

% Add brace showing integer nature
\draw[decorate, decoration={brace, amplitude=4pt, mirror}, gray!50] 
    (9.7,0) -- (9.7,-3*\stepheight) node[midway, right, xshift=3pt, align=left, font=\tiny] 
    {Integer arithmetic\\ maintains precision};

% Add frequency annotation
\node[annotation, text=gray!60, align=left, font=\tiny] at (1.5,-2.4) 
    {Frequencies: $f(A)=5$, $f(B)=3$, $f(C)=2$\\ 
    Total $=10$, $R=2^{16}=65536$};

% Add note about truncation
\node[annotation, text=orange!60!black, font=\tiny, align=left]  at (1.5,-4.8) 
    {Integer division truncation\\ causes slight drift};

\end{tikzpicture}
\caption{Range Coding Integer-Interval Refinement for Message \texttt{BAC}. Real-number equivalents are shown to the right.}
\label{fig:range}
\end{figure}

\subsection{Range Coding Example with Renormalization}

Now we present a more realistic example that demonstrates renormalization in action. We encode a longer message that will cause the range to become small enough to trigger byte output.

Consider the same alphabet $\{A,B,C\}$ with frequencies:

\[
f(A) = 5,\quad f(B) = 3,\quad f(C) = 2,\quad \text{total} = 10.
\]

Cumulative frequencies:

\[
C(A) = 0,\quad C(B) = 5,\quad C(C) = 8.
\]

We now encode a longer message:

\[
\text{MESSAGE} = \texttt{B A C C C C}
\]

\subsubsection*{Step 0: Initialization}
\[
\text{low} = 0,\qquad \text{range} = 2^{32} = 4{,}294{,}967{,}296
\]

\subsubsection*{Step 1: Compute cumulative frequencies}
\[
C(A)=0,\quad C(B)=5,\quad C(C)=8,\quad \text{total}=10
\]

\subsubsection*{Step 2: Encode each symbol in the message}

\textbf{Symbol 1: B}

\begin{align*}
\text{unit} &= \frac{4{,}294{,}967{,}296}{10} = 429{,}496{,}729 \\
\text{low} &= 0 + 429{,}496{,}729 \times 5 = 2{,}147{,}483{,}645 \\
\text{range} &= 429{,}496{,}729 \times 3 = 1{,}288{,}490{,}187
\end{align*}

Renormalization check: $\text{range} = 1{,}288{,}490{,}187 \ge 2^{24} = 16{,}777{,}216$, so continue.

\vspace{2mm}
\textbf{Symbol 2: A}

\begin{align*}
\text{unit} &= \frac{1{,}288{,}490{,}187}{10} = 128{,}849{,}018 \\
\text{low} &= 2{,}147{,}483{,}645 + 128{,}849{,}018 \times 0 = 2{,}147{,}483{,}645 \\
\text{range} &= 128{,}849{,}018 \times 5 = 644{,}245{,}090
\end{align*}

Renormalization check: $\text{range} = 644{,}245{,}090 \ge 16{,}777{,}216$, so continue.

\vspace{2mm}
\textbf{Symbol 3: C}

\begin{align*}
\text{unit} &= \frac{644{,}245{,}090}{10} = 64{,}424{,}509 \\
\text{low} &= 2{,}147{,}483{,}645 + 64{,}424{,}509 \times 8 = 2{,}147{,}483{,}645 + 515{,}396{,}072 = 2{,}662{,}879{,}717 \\
\text{range} &= 64{,}424{,}509 \times 2 = 128{,}849{,}018
\end{align*}

Renormalization check: $\text{range} = 128{,}849{,}018 \ge 16{,}777{,}216$, so continue.

\vspace{2mm}
\textbf{Symbol 4: C}

\begin{align*}
\text{unit} &= \frac{128{,}849{,}018}{10} = 12{,}884{,}901 \\
\text{low} &= 2{,}662{,}879{,}717 + 12{,}884{,}901 \times 8 = 2{,}662{,}879{,}717 + 103{,}079{,}208 = 2{,}765{,}958{,}925 \\
\text{range} &= 12{,}884{,}901 \times 2 = 25{,}769{,}802
\end{align*}

Renormalization check: $\text{range} = 25{,}769{,}802 \ge 16{,}777{,}216$, so continue.

\vspace{2mm}
\textbf{Symbol 5: C}

\begin{align*}
\text{unit} &= \frac{25{,}769{,}802}{10} = 2{,}576{,}980 \\
\text{low} &= 2{,}765{,}958{,}925 + 2{,}576{,}980 \times 8 = 2{,}765{,}958{,}925 + 20{,}615{,}840 = 2{,}786{,}574{,}765 \\
\text{range} &= 2{,}576{,}980 \times 2 = 5{,}153{,}960
\end{align*}

Renormalization check: $\text{range} = 5{,}153{,}960 < 16{,}777{,}216$ — \textbf{renormalization triggered!}

\subsubsection*{Renormalization (after Symbol 5)}

We output bytes until $\text{range} \ge 2^{24}$.

\textbf{Iteration 1:}
\begin{align*}
\text{output byte} &= \text{low} \gg 24 = 2{,}786{,}574{,}765 \gg 24 = 166 \quad (\text{0xA6}) \\
\text{low} &= (\text{low} \ll 8) \mathrel{\&} (2^{32}-1) = 439{,}463{,}680 \\
\text{range} &= \text{range} \ll 8 = 5{,}153{,}960 \times 256 = 1{,}319{,}413{,}760
\end{align*}

Now $\text{range} = 1{,}319{,}413{,}760 \ge 16{,}777{,}216$, so stop renormalization.

Output byte: $\text{0xA6}$ (166).

\vspace{2mm}
\textbf{Symbol 6: C}

\begin{align*}
\text{unit} &= \frac{1{,}319{,}413{,}760}{10} = 131{,}941{,}376 \\
\text{low} &= 439{,}463{,}680 + 131{,}941{,}376 \times 8 = 439{,}463{,}680 + 1{,}055{,}531{,}008 = 1{,}494{,}994{,}688 \\
\text{range} &= 131{,}941{,}376 \times 2 = 263{,}882{,}752
\end{align*}

Renormalization check: $\text{range} = 263{,}882{,}752 \ge 16{,}777{,}216$, so continue.

\subsubsection*{Step 3: Generate the final code}

After encoding all symbols, output the remaining bytes of $\text{low}$.

$\text{low} = 1{,}494{,}994{,}688$ in hexadecimal: $\text{0x591C0000}$

The final output bytes are: $\text{0x59}, \text{0x1C}, \text{0x00}, \text{0x00}$.

\subsubsection*{Complete Encoded Output}

The compressed representation consists of:
\begin{itemize}
    \item The byte output during renormalization: $\text{0xA6}$
    \item The final $\text{low}$ bytes: $\text{0x59}, \text{0x1C}, \text{0x00}, \text{0x00}$
\end{itemize}

Thus the encoded stream is: $[166, 89, 28, 0, 0]$ (5 bytes total).

In hexadecimal: $\text{0xA6 0x59 0x1C 0x00 0x00}$.

\subsubsection*{Encoding State Table}

\begin{center}
\begin{tabular}{c|c|c|c|c}
\hline
\textbf{Step} & \textbf{Symbol} & \textbf{Low} & \textbf{Range} & \textbf{Output} \\ \hline
Start & — & $0$ & $4{,}294{,}967{,}296$ & — \\
1 & B & $2{,}147{,}483{,}645$ & $1{,}288{,}490{,}187$ & — \\
2 & A & $2{,}147{,}483{,}645$ & $644{,}245{,}090$ & — \\
3 & C & $2{,}662{,}879{,}717$ & $128{,}849{,}018$ & — \\
4 & C & $2{,}765{,}958{,}925$ & $25{,}769{,}802$ & — \\
5 & C & $2{,}786{,}574{,}765$ & $5{,}153{,}960$ & — \\
Renorm & — & $439{,}463{,}680$ & $1{,}319{,}413{,}760$ & $166$ \\
6 & C & $1{,}494{,}994{,}688$ & $263{,}882{,}752$ & — \\ \hline
\end{tabular}
\end{center}

\subsubsection*{Why This Sequence Triggers Renormalization}

Four consecutive $C$'s (each with frequency $2/10 = 0.2$) cause the range to shrink by a factor of $0.2$ each time:

\begin{align*}
644{,}245{,}090 \times 0.2 &= 128{,}849{,}018 \\
128{,}849{,}018 \times 0.2 &= 25{,}769{,}802 \\
25{,}769{,}802 \times 0.2 &= 5{,}153{,}960 < 16{,}777{,}216
\end{align*}

This demonstrates how range coding handles interval narrowing through renormalization, outputting bytes as they become fixed.

\subsection{Range Coding: Decoding Algorithm}

Decoding is the inverse process. Given the compressed stream, we reconstruct the original message by determining which symbol's interval contains the current code value.

\begin{tcolorbox}[colback=blue!3,colframe=blue!40!black,title=Algorithm: Range Coding Decoding]
\textbf{Step 0: Initialization} \\
Set $\text{low} = 0$ and $\text{range} = R$ (same as encoder). Read the first 4 bytes from the compressed stream to form the initial $\text{code}$.

\vspace{2mm}
\textbf{Step 1: Compute cumulative frequencies} \\
Same as encoder: $C(s_1) = 0$, $C(s_i) = \sum_{j < i} f(s_j)$, and $\text{total} = \sum_i f(s_i)$.

\vspace{2mm}
\textbf{Step 2: For each symbol to be decoded} \\
For each symbol position in the message (until we have decoded all symbols), perform:
\begin{enumerate}
    \item Compute the offset within the current interval:
    \[
    \text{offset} = \frac{\text{code} - \text{low}}{\text{range}}
    \]
    \item Find the symbol $s$ such that:
    \[
    \frac{C(s)}{\text{total}} \le \text{offset} < \frac{C(s) + f(s)}{\text{total}}
    \]
    (Using integer arithmetic, find $s$ where $C(s) \le \frac{(\text{code} - \text{low}) \times \text{total}}{\text{range}} < C(s) + f(s)$)
    
    \item Output symbol $s$
    
    \item Update $\text{low}$ and $\text{range}$ using the same formulas as the encoder:
    \begin{align*}
    \text{unit} &= \text{range} / \text{total} \\
    \text{low} &= \text{low} + \text{unit} \times C(s) \\
    \text{range} &= \text{unit} \times f(s)
    \end{align*}
    
    \item \textbf{Renormalization:} While $\text{range} < 2^{24}$, read the next byte from the stream and update:
    \begin{align*}
    \text{low} &= (\text{low} \ll 8) \mathrel{\&} (2^{32}-1) \\
    \text{code} &= ((\text{code} \ll 8) \mathrel{\&} (2^{32}-1)) \mid \text{next\_byte} \\
    \text{range} &= \text{range} \ll 8
    \end{align*}
\end{enumerate}
\end{tcolorbox}

\subsection{Range Coding Decoding Example}

We now decode the compressed stream produced in the previous normalization example. The encoded stream was:

\[
[166, 89, 28, 0, 0] \quad \text{in decimal, or} \quad [\text{0xA6}, \text{0x59}, \text{0x1C}, \text{0x00}, \text{0x00}] \text{ in hex}
\]

We know the original message was \texttt{B A C C C C} (6 symbols). We will decode it to verify.

\subsubsection*{Step 0: Initialization}
\[
\text{low} = 0,\qquad \text{range} = 2^{32} = 4{,}294{,}967{,}296
\]

Read the first 4 bytes to form the initial code:
\[
\text{code} = 0\text{xA6591C00} = 2{,}789{,}021{,}696
\]

\subsubsection*{Step 1: Compute cumulative frequencies}
\[
f(A)=5,\quad f(B)=3,\quad f(C)=2,\quad \text{total}=10
\]
\[
C(A)=0,\quad C(B)=5,\quad C(C)=8
\]

\subsubsection*{Step 2: Decode each symbol}

\textbf{Symbol 1:}

Compute offset:
\[
\text{offset} = \frac{\text{code} - \text{low}}{\text{range}} = \frac{2{,}789{,}021{,}696}{4{,}294{,}967{,}296} \approx 0.6494
\]

Find symbol containing 0.6494:
\begin{itemize}
    \item A: $[0, 0.5)$ — no
    \item B: $[0.5, 0.8)$ — yes, contains 0.6494
    \item C: $[0.8, 1.0)$ — no
\end{itemize}

Decoded symbol: \textbf{B}

Update using B ($C(B)=5$, $f(B)=3$):
\begin{align*}
\text{unit} &= \frac{4{,}294{,}967{,}296}{10} = 429{,}496{,}729 \\
\text{low} &= 0 + 429{,}496{,}729 \times 5 = 2{,}147{,}483{,}645 \\
\text{range} &= 429{,}496{,}729 \times 3 = 1{,}288{,}490{,}187
\end{align*}

Renormalization check: $\text{range} = 1{,}288{,}490{,}187 \ge 16{,}777{,}216$, so continue.

\vspace{2mm}
\textbf{Symbol 2:}

Compute offset:
\[
\text{offset} = \frac{2{,}789{,}021{,}696 - 2{,}147{,}483{,}645}{1{,}288{,}490{,}187} = \frac{641{,}538{,}051}{1{,}288{,}490{,}187} \approx 0.4980
\]

Find symbol containing 0.4980:
\begin{itemize}
    \item A: $[0, 0.5)$ — yes, contains 0.4980
    \item B: $[0.5, 0.8)$ — no
    \item C: $[0.8, 1.0)$ — no
\end{itemize}

Decoded symbol: \textbf{A}

Update using A ($C(A)=0$, $f(A)=5$):
\begin{align*}
\text{unit} &= \frac{1{,}288{,}490{,}187}{10} = 128{,}849{,}018 \\
\text{low} &= 2{,}147{,}483{,}645 + 128{,}849{,}018 \times 0 = 2{,}147{,}483{,}645 \\
\text{range} &= 128{,}849{,}018 \times 5 = 644{,}245{,}090
\end{align*}

Renormalization check: $\text{range} = 644{,}245{,}090 \ge 16{,}777{,}216$, so continue.

\vspace{2mm}
\textbf{Symbol 3:}

Compute offset:
\[
\text{offset} = \frac{2{,}789{,}021{,}696 - 2{,}147{,}483{,}645}{644{,}245{,}090} = \frac{641{,}538{,}051}{644{,}245{,}090} \approx 0.9958
\]

Find symbol containing 0.9958:
\begin{itemize}
    \item A: $[0, 0.5)$ — no
    \item B: $[0.5, 0.8)$ — no
    \item C: $[0.8, 1.0)$ — yes, contains 0.9958
\end{itemize}

Decoded symbol: \textbf{C}

Update using C ($C(C)=8$, $f(C)=2$):
\begin{align*}
\text{unit} &= \frac{644{,}245{,}090}{10} = 64{,}424{,}509 \\
\text{low} &= 2{,}147{,}483{,}645 + 64{,}424{,}509 \times 8 = 2{,}147{,}483{,}645 + 515{,}396{,}072 = 2{,}662{,}879{,}717 \\
\text{range} &= 64{,}424{,}509 \times 2 = 128{,}849{,}018
\end{align*}

Renormalization check: $\text{range} = 128{,}849{,}018 \ge 16{,}777{,}216$, so continue.

\vspace{2mm}
\textbf{Symbol 4:}

Compute offset:
\[
\text{offset} = \frac{2{,}789{,}021{,}696 - 2{,}662{,}879{,}717}{128{,}849{,}018} = \frac{126{,}141{,}979}{128{,}849{,}018} \approx 0.9790
\]

Find symbol containing 0.9790:
\begin{itemize}
    \item A: $[0, 0.5)$ — no
    \item B: $[0.5, 0.8)$ — no
    \item C: $[0.8, 1.0)$ — yes
\end{itemize}

Decoded symbol: \textbf{C}

Update using C ($C(C)=8$, $f(C)=2$):
\begin{align*}
\text{unit} &= \frac{128{,}849{,}018}{10} = 12{,}884{,}901 \\
\text{low} &= 2{,}662{,}879{,}717 + 12{,}884{,}901 \times 8 = 2{,}662{,}879{,}717 + 103{,}079{,}208 = 2{,}765{,}958{,}925 \\
\text{range} &= 12{,}884{,}901 \times 2 = 25{,}769{,}802
\end{align*}

Renormalization check: $\text{range} = 25{,}769{,}802 \ge 16{,}777{,}216$, so continue.

\vspace{2mm}
\textbf{Symbol 5:}

Compute offset:
\[
\text{offset} = \frac{2{,}789{,}021{,}696 - 2{,}765{,}958{,}925}{25{,}769{,}802} = \frac{23{,}062{,}771}{25{,}769{,}802} \approx 0.8950
\]

Find symbol containing 0.8950:
\begin{itemize}
    \item A: $[0, 0.5)$ — no
    \item B: $[0.5, 0.8)$ — no
    \item C: $[0.8, 1.0)$ — yes
\end{itemize}

Decoded symbol: \textbf{C}

Update using C ($C(C)=8$, $f(C)=2$):
\begin{align*}
\text{unit} &= \frac{25{,}769{,}802}{10} = 2{,}576{,}980 \\
\text{low} &= 2{,}765{,}958{,}925 + 2{,}576{,}980 \times 8 = 2{,}765{,}958{,}925 + 20{,}615{,}840 = 2{,}786{,}574{,}765 \\
\text{range} &= 2{,}576{,}980 \times 2 = 5{,}153{,}960
\end{align*}

Renormalization check: $\text{range} = 5{,}153{,}960 < 16{,}777{,}216$ — \textbf{renormalization triggered!}

\subsubsection*{Renormalization During Decoding}

We read the next byte from the stream (the fifth byte: $\text{0x00}$):

\begin{align*}
\text{low} &= (\text{low} \ll 8) \mathrel{\&} (2^{32}-1) = 439{,}463{,}680 \\
\text{code} &= ((\text{code} \ll 8) \mathrel{\&} (2^{32}-1)) \mid 0 = 1{,}494{,}994{,}688 \\
\text{range} &= \text{range} \ll 8 = 1{,}319{,}413{,}760
\end{align*}

Now $\text{range} = 1{,}319{,}413{,}760 \ge 16{,}777{,}216$, so stop renormalization.

\vspace{2mm}
\textbf{Symbol 6:}

Compute offset:
\[
\text{offset} = \frac{1{,}494{,}994{,}688 - 439{,}463{,}680}{1{,}319{,}413{,}760} = \frac{1{,}055{,}531{,}008}{1{,}319{,}413{,}760} = 0.8
\]

Find symbol containing 0.8:
\begin{itemize}
    \item A: $[0, 0.5)$ — no
    \item B: $[0.5, 0.8)$ — note: B is $[0.5, 0.8)$, which \textbf{excludes} 0.8
    \item C: $[0.8, 1.0)$ — yes, contains 0.8 (the left endpoint is included)
\end{itemize}

Decoded symbol: \textbf{C}

\subsubsection*{Decoding Complete}

We have recovered the original sequence:
\[
\boxed{\texttt{B A C C C C}}
\]

\subsubsection*{Decoding State Table}

\begin{center}
\begin{tabular}{c|c|c|c|c}
\hline
\textbf{Step} & \textbf{Code} & \textbf{Low} & \textbf{Range} & \textbf{Symbol} \\ \hline
Start & $2{,}789{,}021{,}696$ & $0$ & $4{,}294{,}967{,}296$ & — \\
1 & $2{,}789{,}021{,}696$ & $2{,}147{,}483{,}645$ & $1{,}288{,}490{,}187$ & B \\
2 & $2{,}789{,}021{,}696$ & $2{,}147{,}483{,}645$ & $644{,}245{,}090$ & A \\
3 & $2{,}789{,}021{,}696$ & $2{,}662{,}879{,}717$ & $128{,}849{,}018$ & C \\
4 & $2{,}789{,}021{,}696$ & $2{,}765{,}958{,}925$ & $25{,}769{,}802$ & C \\
5 & $2{,}789{,}021{,}696$ & $2{,}786{,}574{,}765$ & $5{,}153{,}960$ & C \\
Renorm & $1{,}494{,}994{,}688$ & $439{,}463{,}680$ & $1{,}319{,}413{,}760$ & — \\
6 & $1{,}494{,}994{,}688$ & $439{,}463{,}680$ & $1{,}319{,}413{,}760$ & C \\ \hline
\end{tabular}
\end{center}

\subsubsection*{Key Observations}

\begin{enumerate}
    \item \textbf{Decoding mirrors encoding:} The decoder uses the same cumulative frequencies and update formulas as the encoder.
    
    \item \textbf{Renormalization symmetry:} The encoder outputs a byte when $\text{range} < 2^{24}$; the decoder reads that same byte when $\text{range} < 2^{24}$ and incorporates it into $\text{code}$. Both update $\text{low}$ and $\text{range}$ identically.
    
    \item \textbf{Offset calculation:} The offset determines which symbol's interval contains the current code value. The symbol is found by comparing the scaled offset with cumulative frequencies.
    
    \item \textbf{Precision maintenance:} Renormalization ensures that $\text{range}$ remains large enough to maintain precision throughout the decoding process.
\end{enumerate}

\subsection{Direct Comparison: Arithmetic vs. Range Coding}

The following table summarizes the parallel between the two methods:

\begin{center}
\begin{tabular}{l | l | l}
\toprule
\textbf{Concept} & \textbf{Arithmetic Coding} & \textbf{Range Coding} \\
\midrule
Representation & Real numbers in $[0,1)$ & Integers in $[0, R)$ \\ \hline
Initial state & $\text{low}=0$, $\text{high}=1$ & $\text{low}=0$, $\text{range}=R$ \\ \hline
Unit size & $\text{range} = \text{high}-\text{low}$ & $\text{unit} = \text{range} / \text{total}$ \\ \hline
Low update & $\text{low} + \text{range} \times C(s)$ & $\text{low} + \text{unit} \times C(s)$ \\ \hline
Range update & $\text{range} \times p(s)$ & $\text{unit} \times f(s)$ \\ \hline
Precision handling & Floating-point normalization & Integer renormalization \\ \hline
Final code & Midpoint of final interval & Integer from final interval \\ \hline
\bottomrule
\end{tabular}
\end{center}

\subsection{Summary}

\begin{itemize}
    \item Arithmetic coding uses real-number intervals and is mathematically elegant.
    \item Range coding uses integer intervals but produces the same final interval.
    \item Range coding is faster in practice because it avoids floating-point arithmetic and uses only integer operations.
    \item Both achieve compression close to the theoretical entropy limit.
    \item The two methods are mathematically equivalent; range coding is simply an integer-based implementation of arithmetic coding.
    \item Renormalization in range coding is integrated into the symbol encoding loop, outputting bytes when the range becomes too small.
    \item Decoding mirrors encoding, using the same cumulative frequencies and update formulas to reconstruct the original message.
\end{itemize}

\subsection{Exercises for Students}

\begin{enumerate}
    \item Encode the message \texttt{CAB} using arithmetic coding with the same probabilities. Show each step.
    \item Encode the same message \texttt{CAB} using range coding with $R=65536$ and compare the final intervals.
    \item Explain why range coding is more suitable for hardware implementation than arithmetic coding.
    \item What would happen if we used a smaller initial range, such as $R=1000$, for range coding? How would it affect compression quality?
    \item Write pseudocode for an arithmetic coding encoder and a range coding encoder. Identify the key differences.
    \item Using the figures provided, explain how the interval refinement process differs between arithmetic coding and range coding.
    \item Why is the update order (first high, then low) important in arithmetic coding?
    \item What is the purpose of renormalization in range coding? In the normalization example, why did we need to renormalize after encoding the fifth symbol?
    \item Modify the normalization example to encode the message \texttt{B A C C} instead. Would renormalization still be triggered? Why or why not?
    \item Using the decoding algorithm, decode the compressed output from Exercise 2 (the \texttt{CAB} message) to verify you recover the original message.
\end{enumerate}

\end{document}
