\section{Transform Coding and JPEG: The DCT Revolution}

\subsection{Learning Objectives}

By the end of this lecture, you will be able to:
\begin{enumerate}[leftmargin=*, nosep]
    \item Explain why transform coding is necessary for efficient lossy compression.
    \item Describe the energy-compaction property of the DCT and state why it approximates the optimal KL transform for natural images.
    \item Execute the complete JPEG pipeline on an $8\times8$ block: level-shift $\to$ 2D DCT $\to$ quantisation $\to$ zig-zag scan $\to$ RLE $\to$ Huffman.
    \item Quantify the rate–distortion trade-off controlled by the JPEG quality factor.
    \item Identify the physical cause of blocking, ringing, banding, and colour-bleeding artefacts.
    \item Connect JPEG's entropy-coding stage to Huffman coding and RLE from earlier lectures.
\end{enumerate}

\subsection{Motivation: Why Transform Coding?}

\subsubsection{Spatial Redundancy in Natural Images}

Raw pixel values in natural images are \emph{highly correlated}: a pixel is almost always very similar to its neighbours. Table~\ref{tab:rawblock} shows a representative $8\times8$ luminance block taken from a photograph of a clear sky. Notice how the values change slowly and smoothly : there are no sudden jumps.

\begin{table}[H]
\centering
\caption{Raw $8\times8$ luminance block (pixel values 0–255). Slow spatial variation is evident.}
\label{tab:rawblock}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\rowcolor{LightBlue!20}
142 & 148 & 155 & 160 & 158 & 150 & 145 & 140 \\
\hline
140 & 146 & 153 & 158 & 156 & 148 & 143 & 138 \\
\hline
\rowcolor{LightGray!10}
138 & 144 & 151 & 156 & 154 & 146 & 141 & 136 \\
\hline
135 & 141 & 148 & 153 & 151 & 143 & 138 & 133 \\
\hline
\rowcolor{LightGray!10}
132 & 138 & 145 & 150 & 148 & 140 & 135 & 130 \\
\hline
130 & 136 & 143 & 148 & 146 & 138 & 133 & 128 \\
\hline
\rowcolor{LightGray!10}
128 & 134 & 141 & 146 & 144 & 136 & 131 & 126 \\
\hline
125 & 131 & 138 & 143 & 141 & 133 & 128 & 123 \\
\hline
\end{tabular}
\end{table}

\begin{importantbox}
\textbf{The Core Problem:} Huffman coding assigns short codes to \emph{frequent symbols}. But here every pixel value appears at most once : the symbol frequencies are nearly flat, so Huffman gains almost nothing. The redundancy is \emph{structural}, encoded in smooth spatial transitions, not in symbol repetition. We need a different representation to expose it.
\end{importantbox}

\subsubsection{An Intuitive Analogy: Reorganising a Library}

Imagine you have a library where books are arranged randomly. To find a specific topic, you must search everywhere : this is like raw pixels. Now imagine reorganising the library so that all books on the same topic are on the same shelf. The information (the books) hasn't changed, but the \emph{organisation} makes it much easier to find what you need.

\textbf{Transform coding does exactly this:} it reorganises the information in an image block so that the most important parts (the "topics") are grouped together in a few coefficients, while the less important details (the "random noise") become small and can be discarded.

\subsubsection{The Transform-Coding Idea}

\begin{definitionbox}
\textbf{Transform Coding:} Apply a reversible linear transform $T$ to convert a block of $N$ correlated samples into $N$ \textbf{decorrelated coefficients} such that most of the signal energy concentrates into a small number of coefficients. The transform itself is \emph{lossless}; loss is introduced later, during quantisation.
\end{definitionbox}

The conceptual chain is:

\vspace{6pt}
\begin{center}
\begin{tikzpicture}[
    node distance = 1.2cm,
    box/.style = {
        draw,
        rounded corners=4pt,
        minimum width=2.6cm,
        minimum height=0.9cm,
        align=center,
        font=\small,
        text=black,
        fill opacity=1
    },
    lossless/.style = {box, fill=blue!15, draw=blue!70!black},
    lossy/.style    = {box, fill=red!15,  draw=red!70!black},
    arr/.style      = {-{Stealth[length=5pt]}, thick}
  ]

  \node[lossless] (T) {Transform\\(DCT)};
  \node[lossy, right=of T] (Q) {Quantisation};
  \node[lossless, right=of Q] (E) {Entropy\\Coding};

  \draw[arr] (T) -- (Q);
  \draw[arr] (Q) -- (E);

  \node[below=0.2cm of T, font=\scriptsize\itshape, text=blue!70!black] {lossless};
  \node[below=0.2cm of Q, font=\scriptsize\bfseries, text=red!70!black] {LOSSY};
  \node[below=0.2cm of E, font=\scriptsize\itshape, text=blue!70!black] {lossless};

\end{tikzpicture}
\end{center}
\vspace{4pt}

This three-stage structure is \textbf{universal}: JPEG (images), MP3/AAC (audio), H.264/H.265 (video), and HEIC all follow it.

\begin{importantbox}
\textbf{Key Insight:} The transform stage does NOT compress anything. It simply rearranges the information. The actual compression happens during quantisation, where we deliberately discard information that the human eye is unlikely to notice.
\end{importantbox}

\subsection{The Discrete Cosine Transform}

\subsubsection{Intuition: Change of Basis}

Think of an $8\times8$ image block as a point in a 64-dimensional space. The standard basis for this space is the set of 64 matrices that have a 1 in one position and 0 everywhere else : each basis vector represents a single pixel.

The DCT provides a different basis for the same space. Instead of "single pixel" basis vectors, the DCT basis vectors are patterns of cosines at different frequencies. Any image block can be written as a weighted sum of these 64 fixed $8\times8$ \textbf{basis patterns} $B_{u,v}(x,y)$:

\[
p(x,y) = \sum_{u=0}^{7}\sum_{v=0}^{7} D(u,v)\,B_{u,v}(x,y).
\]

The coefficients $D(u,v)$ are the \emph{coordinates} of the block in the DCT basis. If a coefficient is large, it means the image block strongly resembles that particular frequency pattern.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.85]
  % 8x8 grid with brightness decreasing toward bottom-right
  \foreach \u in {0,...,7}{
    \foreach \v in {0,...,7}{
      \pgfmathtruncatemacro{\bval}{95 - (\u+\v)*6}
      \pgfmathtruncatemacro{\bval}{max(\bval,30)}  % Minimum 30% to keep text visible
      \fill[gray!\bval!white] (\u*0.82, \v*0.82) rectangle ++(0.80,0.80);
      \draw[gray!40, line width=0.3pt] (\u*0.82, \v*0.82) rectangle ++(0.80,0.80);
    }
  }
  % Highlight DC cell with orange border
  \draw[orange, line width=2pt] (0,7*0.82) rectangle ++(0.80,0.80);
  \node[font=\tiny\bfseries, text=black] at (0.40, 7*0.82+0.40) {DC};
  % Add frequency labels with white background for visibility
  \node[below, font=\footnotesize, fill=white, fill opacity=0.7, text opacity=1, inner sep=1pt] at (3.28,-0.25) {horizontal frequency $u \longrightarrow$};
  \node[left, rotate=90, font=\footnotesize, fill=white, fill opacity=0.7, text opacity=1, inner sep=1pt] at (-0.3, 3.28) {vertical frequency $v \longrightarrow$};
\end{tikzpicture}
\caption{Schematic of the 64 DCT basis functions for an $8\times8$ block. The top-left cell (orange border) is the \textbf{DC} basis : a uniform block representing average brightness. Moving right increases horizontal frequency; moving down increases vertical frequency. Natural images concentrate energy in the top-left (low-frequency) region.}
\label{fig:basis}
\end{figure}

\subsubsection{The 2D DCT Formula}

\begin{definitionbox}
\textbf{2D DCT (Type-II):} For an $8\times8$ block $p(x,y)$, $x,y\in\{0,\ldots,7\}$:
\begin{equation}
\boxed{
  D(u,v) = \frac{C(u)\,C(v)}{4}
  \sum_{x=0}^{7}\sum_{y=0}^{7}
  p(x,y)\,
  \cos\!\left(\frac{(2x+1)\,u\,\pi}{16}\right)
  \cos\!\left(\frac{(2y+1)\,v\,\pi}{16}\right)
}
\label{eq:dct}
\end{equation}
where
\begin{equation}
  C(k) = \begin{cases} 1/\sqrt{2} & k=0, \\ 1 & k > 0. \end{cases}
\end{equation}
\end{definitionbox}

\paragraph{What does this formula actually compute?}

For each output coefficient $D(u,v)$, we are:
\begin{enumerate}[nosep]
    \item Taking every input pixel $p(x,y)$
    \item Multiplying it by two cosine values (one for the horizontal frequency $u$, one for the vertical frequency $v$)
    \item Summing all these products together
    \item Scaling the result by $\frac{C(u)C(v)}{4}$
\end{enumerate}

The cosine values range from $-1$ to $+1$. When $u=0$ (the lowest horizontal frequency), $\cos(\frac{(2x+1)\cdot0\cdot\pi}{16}) = \cos(0) = 1$ for all $x$. This is why the DC coefficient (where $u=0,v=0$) simply sums all the pixels: it captures the average brightness.

\paragraph{Inverse DCT (IDCT):} The transform is orthogonal, so the inverse uses the same cosine kernels with roles swapped:
\begin{equation}
  \hat{p}(x,y) = \frac{1}{4}
  \sum_{u=0}^{7}\sum_{v=0}^{7}
  C(u)\,C(v)\,D(u,v)\,
  \cos\!\left(\frac{(2x+1)\,u\,\pi}{16}\right)
  \cos\!\left(\frac{(2y+1)\,v\,\pi}{16}\right).
  \label{eq:idct}
\end{equation}
After exact (unquantised) DCT$\to$IDCT, $\hat{p}=p$ to floating-point precision.

\subsubsection{Separability and Computational Cost}

Because the cosine kernel factorises as $\cos(\cdots u\cdots)\cos(\cdots v\cdots)$, the 2D DCT is \textbf{separable}. We can rewrite Eq.~\eqref{eq:dct} as:

\begin{equation}
\boxed{
  D(u,v) = \frac{C(u)C(v)}{4} \sum_{x=0}^{7} \left[
    \cos\!\left(\frac{(2x+1)u\pi}{16}\right)
    \underbrace{\left( \sum_{y=0}^{7} p(x,y) \cos\!\left(\frac{(2y+1)v\pi}{16}\right) \right)}_{T(x,v)}
  \right]
}
\label{eq:separable}
\end{equation}

where we define the \textbf{intermediate coefficient} $T(x,v)$ as the 1D DCT of row $x$ at vertical frequency $v$ (without scaling):

\begin{equation}
  T(x,v) \triangleq \sum_{y=0}^{7} p(x,y) \cos\!\left(\frac{(2y+1)v\pi}{16}\right), \qquad x,v \in \{0,\ldots,7\}
\end{equation}

This $T(x,v)$ is the raw sum-of-products before any scaling. The complete 1D row transform would be $\frac{C(v)}{2} T(x,v)$, but we delay scaling until the final step for efficiency.

\paragraph{The two-pass separable algorithm:}

\begin{enumerate}[nosep]
    \item \textbf{Row pass (1D DCT on each row):} 
    For each row index $x$ and each frequency $v$, compute $T(x,v)$ using the formula above. This produces an $8\times8$ intermediate block whose $x$-th row contains the unscaled 1D DCT of the $x$-th input row.
    
    \item \textbf{Column pass (1D DCT on each column of $T$):} 
    For each frequency $v$ (which indexes columns of $T$) and each output frequency $u$, apply the 1D DCT to the column vector $T(0..7, v)$:
    \begin{equation}
      D(u,v) = \frac{C(u)C(v)}{4} \sum_{x=0}^{7} T(x,v) \cos\!\left(\frac{(2x+1)u\pi}{16}\right)
    \end{equation}
    The scaling factors $\frac{C(u)C(v)}{4}$ are applied only once at the end.
\end{enumerate}

\paragraph{Matrix formulation:}

The separable nature is even clearer in matrix form:
\begin{equation}
  \mathbf{D} = \mathbf{C} \cdot \mathbf{P} \cdot \mathbf{C}^{\mathsf{T}}
\end{equation}
where:
\begin{itemize}[nosep]
    \item $\mathbf{P}$ is the $8\times8$ pixel block ($P_{x,y} = p(x,y)$)
    \item $\mathbf{C}$ is the $8\times8$ DCT basis matrix with entries $C_{u,x} = \frac{C(u)}{2}\cos\!\left(\frac{(2x+1)u\pi}{16}\right)$
    \item $\mathbf{D}$ is the $8\times8$ DCT coefficient matrix ($D_{u,v} = D(u,v)$)
\end{itemize}
The product $\mathbf{C} \cdot \mathbf{P}$ performs the row transforms (scaled), and multiplying by $\mathbf{C}^{\mathsf{T}}$ on the right performs the column transforms.

\paragraph{Computational Complexity Comparison}

Let $N=8$ for JPEG blocks. Count the number of multiply-add operations required:

\begin{table}[h]
\centering
\begin{tabular}{|l|l|c|}
\hline
\textbf{Method} & \textbf{Operation count formula} & \textbf{For $N=8$} \\ \hline
Naive 2D (Eq.~\ref{eq:dct}) & $N^2$ coefficients $\times$ $N^2$ terms & $4096$ \\ \hline
Separable (Eq.~\ref{eq:separable}) & $2 \times (N \times N^2)$ & $1024$ \\ \hline
Fast DCT (FFT-like) & $2 \times (N \times \frac{N}{2}\log_2 N)$ & $\approx 192$ \\ \hline
\end{tabular}
\caption{Operation counts and asymptotic complexity for one $8\times8$ block. Naive: $O(N^4)$, separable: $O(N^3)$, fast: $O(N^2\log N)$.}
\label{tab:complexity}
\end{table}



\begin{importantbox}
\textbf{Why Separability Matters:} Direct 2D summation in Eq.~\eqref{eq:dct} costs $O(N^4)$ multiplications for an $N\times N$ block. Separability reduces this to $O(N^3)$, and the fast DCT algorithm (analogous to the FFT) reduces it further to $O(N^2\log N)$.

For $N=8$:
\begin{itemize}[nosep]
    \item Naive 2D: $4096$ multiplications
    \item Fast separable DCT: $\approx 64$ multiplications (using optimized algorithms)
    \item Speedup: $4096 / 64 = \mathbf{64\times}$
\end{itemize}
This is the difference between a practical encoder and an impractical one!
\end{importantbox}

\paragraph{Where does the $64$ come from?}

The fast 1D DCT of length 8 requires approximately $\frac{N}{2}\log_2 N = 4 \times 3 = 12$ multiplications by exploiting cosine symmetries (many basis functions share the same magnitude). For the full 2D transform:
\begin{itemize}[nosep]
    \item 8 row transforms: $8 \times 12 = 96$ multiplications
    \item 8 column transforms: $8 \times 12 = 96$ multiplications
    \item Total: $\approx 192$ multiplications
\end{itemize}
The $64\times$ speedup figure (4096 vs 64) assumes a highly optimized fast DCT that further prunes operations through combined row-column planning and coefficient precomputation. Even the conservative $192$ operations represent a $21\times$ speedup over naive $4096$ — both are massive gains that make real-time JPEG encoding feasible.


\begin{examplebox}
\textbf{A Complete Worked Example: 2$\times$2 DCT Using the Separable Formula}

To truly understand the 2D DCT, we work through a \textbf{tiny $2\times2$ block}
by hand using the \textbf{separable (two-pass) approach}. This matches how real
JPEG encoders work: first transform all rows, then all columns.

\medskip
\noindent
\textbf{Input pixel block $\mathbf{P}$ (a simple 2$\times$2 gradient):}
\[
\mathbf{P} = \begin{bmatrix}
p(0,0) & p(0,1) \\
p(1,0) & p(1,1)
\end{bmatrix}
= \begin{bmatrix}
100 & 120 \\
140 & 160
\end{bmatrix}
\]
This block has a clear increasing gradient from top‑left to bottom‑right.

\medskip
\noindent
\textbf{\large Step 1: Understand the 1D DCT Formula (Length 2)}

For a single row of length $N=2$, the 1D DCT is:
\[
D_{1D}(v) = \frac{C(v)}{2} \sum_{y=0}^{1} p(y) \cos\!\left(\frac{(2y+1)v\pi}{4}\right)
\]
where $C(0)=1/\sqrt{2} \approx 0.7071$ and $C(1)=1$.

Precompute the cosine values:
\begin{center}
\begin{tabular}{c|c|c}
$y$ & $v=0$: $\cos(0)$ & $v=1$: $\cos\!\left(\frac{(2y+1)\pi}{4}\right)$ \\ \hline
0 & $1$ & $\cos(45^\circ)=0.7071$ \\
1 & $1$ & $\cos(135^\circ)=-0.7071$ \\
\end{tabular}
\end{center}

\medskip
\noindent
\textbf{\large Step 2: First Pass --- Apply 1D DCT to Each Row}

We compute $T(x,v)$, the \emph{unscaled} 1D DCT of row $x$ at frequency $v$
(the scaling factor $\frac{C(v)}{2}$ is deferred to the final step):
\[
T(x,v) = \sum_{y=0}^{1} p(x,y) \cos\!\left(\frac{(2y+1)v\pi}{4}\right)
\]

\noindent
\textbf{Row 0:} $p(0,0)=100$, $p(0,1)=120$

\begin{align*}
T(0,0) &= 100(1) + 120(1) = 220 \\
T(0,1) &= 100(0.7071) + 120(-0.7071) = 70.71 - 84.85 = -14.14
\end{align*}

\noindent
\textbf{Row 1:} $p(1,0)=140$, $p(1,1)=160$

\begin{align*}
T(1,0) &= 140(1) + 160(1) = 300 \\
T(1,1) &= 140(0.7071) + 160(-0.7071) = 98.99 - 113.14 = -14.14
\end{align*}

\noindent
The intermediate block $T(x,v)$ (rows $x$, columns $v$) is:
\[
\mathbf{T} = \begin{bmatrix}
220 & -14.14 \\
300 & -14.14
\end{bmatrix}
\]
Note that $T(0,1)$ and $T(1,1)$ are \emph{identical}: both rows have the same
difference $160-140 = 140-120 = 20$, scaled by $0.7071$, so both yield
$-14.14$.

\medskip
\noindent
\textbf{\large Step 3: Second Pass --- Apply 1D DCT to Each Column of $\mathbf{T}$}

For each column $v$, treat it as a vector and apply the unscaled 1D DCT again.
The combined scaling factor $\frac{C(u)C(v)}{4}$ will be applied in Step~4.

\[
S(u,v) = \sum_{x=0}^{1} T(x,v) \cos\!\left(\frac{(2x+1)u\pi}{4}\right)
\]

\noindent
\textbf{Column $v=0$:} $T(0,0)=220$, $T(1,0)=300$

\begin{align*}
S(0,0) &= 220(1) + 300(1) = 520 \\
S(1,0) &= 220(0.7071) + 300(-0.7071) = 155.56 - 212.13 = -56.57
\end{align*}

\noindent
\textbf{Column $v=1$:} $T(0,1)=-14.14$, $T(1,1)=-14.14$

\begin{align*}
S(0,1) &= (-14.14)(1) + (-14.14)(1) = -28.28 \\[4pt]
S(1,1) &= (-14.14)(0.7071) + (-14.14)(-0.7071) \\
       &= \underbrace{(-14.14)(0.7071)}_{-10.00} + \underbrace{(-14.14)(-0.7071)}_{+10.00} = 0
\end{align*}

$S(1,1)=0$ \emph{exactly} because $T(0,1)=T(1,1)$: the two terms cancel perfectly.

The unscaled coefficient matrix is:
\[
\mathbf{S} = \begin{bmatrix}
520 & -28.28 \\
-56.57 & 0
\end{bmatrix}
\]

\medskip
\noindent
\textbf{\large Step 4: Apply Final Scaling}

The scaling factor for each coefficient is $\frac{C(u)C(v)}{4}$, combining the
$\frac{C(u)}{2}$ from the column pass and $\frac{C(v)}{2}$ from the row pass.

\begin{align*}
\frac{C(0)C(0)}{4} &= \frac{0.7071 \times 0.7071}{4} = \frac{0.5}{4} = 0.125\\
\frac{C(0)C(1)}{4} &= \frac{C(1)C(0)}{4} = \frac{0.7071 \times 1}{4} \approx 0.17678\\
\frac{C(1)C(1)}{4} &= \frac{1}{4} = 0.25
\end{align*}

Applying to $\mathbf{S}$:

\begin{align*}
D(0,0) &= 0.125   \times 520   &&= 65 \\
D(0,1) &= 0.17678 \times (-28.28) &&= -5.00 \\
D(1,0) &= 0.17678 \times (-56.57) &&= -10.00 \\
D(1,1) &= 0.25    \times 0      &&= 0
\end{align*}

Thus:
\[
\boxed{\mathbf{D} = \begin{bmatrix}
65 & -5 \\
-10 & 0
\end{bmatrix}}
\]

\medskip
\noindent
\textbf{\large Verification Using the Direct 2D Summation Formula}

The direct 2D formula is:
\[
D(u,v) = \frac{C(u)C(v)}{4} \sum_{x=0}^{1}\sum_{y=0}^{1}
p(x,y)\cos\!\left(\frac{(2x+1)u\pi}{4}\right)\cos\!\left(\frac{(2y+1)v\pi}{4}\right)
\]
This is simply the separable formula written out in full, so the scaling is
$\frac{C(u)C(v)}{4}$---identical to what we used above.

Computing $D(0,0)$ directly:
\[
D(0,0) = 0.125 \times (100 + 120 + 140 + 160) \times 1 \times 1
        = 0.125 \times 520 = 65
\]
All four coefficients match exactly, confirming that the separable two-pass
method and the direct formula are equivalent.

\medskip
\noindent
\textbf{\large Connection to Matrix Form}

The separable two-pass algorithm is exactly equivalent to:
\[
\mathbf{D} = \mathbf{C} \cdot \mathbf{P} \cdot \mathbf{C}^{\mathsf{T}}
\]
where $\mathbf{C}$ is the DCT basis matrix with entries
$C_{u,x} = \frac{C(u)}{2}\cos\!\left(\frac{(2x+1)u\pi}{4}\right)$.

For the $2\times2$ case:
\[
\mathbf{C} = \begin{bmatrix}
0.35355 & 0.35355 \\
0.35355 & -0.35355
\end{bmatrix}
\]

Then $\mathbf{C}\mathbf{P}\mathbf{C}^{\mathsf{T}}$ produces exactly the same $\mathbf{D}$:
\[
\begin{bmatrix}
0.35355 & 0.35355 \\
0.35355 & -0.35355
\end{bmatrix}
\begin{bmatrix}
100 & 120 \\
140 & 160
\end{bmatrix}
\begin{bmatrix}
0.35355 & 0.35355 \\
0.35355 & -0.35355
\end{bmatrix}
=
\begin{bmatrix}
65 & -5 \\
-10 & 0
\end{bmatrix}
\]

\medskip
\noindent
\textbf{\large What We Learned}

\begin{enumerate}[nosep]
    \item The \textbf{separable formula} reveals the true computational structure:
          first transform all rows, then all columns.
    \item The \textbf{matrix form} $\mathbf{D} = \mathbf{C}\mathbf{P}\mathbf{C}^{\mathsf{T}}$
          is a compact representation of the same two-pass process.
    \item The DC coefficient $D(0,0)$ captures the average brightness, scaled by
          $\frac{C(0)^2}{4} \times N^2 = \frac{1}{2}$ in this $2\times2$ case.
    \item AC coefficients are small when the block is smooth (our gradient gave
          $D(0,1)=-5$, $D(1,0)=-10$).
    \item $D(1,1)=0$ \emph{exactly} because the checkerboard cosine pattern
          is orthogonal to the uniform gradient: the two rows of $\mathbf{T}$
          have identical entries in column~1, so they cancel in the column pass.
    \item The direct 2D formula and the separable formula use the \emph{same}
          scaling $\frac{C(u)C(v)}{4}$ and always agree; they are two ways
          of writing the same computation.
\end{enumerate}

\begin{center}
\boxed{\text{2D DCT = 1D DCT on rows, then 1D DCT on columns}}
\end{center}

This concrete arithmetic removes the mystery of the DCT and makes the abstract
formulas tangible.
\end{examplebox}


\subsubsection{Why DCT, Not DFT?}

The Discrete Fourier Transform (DFT) requires \emph{complex} coefficients and implicitly assumes the signal repeats periodically. Pixel blocks do not: abrupt discontinuities at block boundaries would introduce large high-frequency DFT coefficients, hurting compression.

The DCT implicitly treats the block as if it were reflected (even-extended) at its boundaries:
\[
\text{Original block: } a, b, c, d \quad \text{Reflected: } a, b, c, d, d, c, b, a
\]
This gives a much smoother periodic signal and therefore better energy compaction.

\begin{definitionbox}
\textbf{Why the DCT is (Nearly) Optimal:} The theoretically optimal linear decorrelating transform for a stationary signal is the \textbf{Karhunen--Lo\`eve Transform (KLT)}, which diagonalises the signal covariance matrix. For a first-order Markov model with correlation coefficient $\rho\to1$ (i.e., slowly varying signals, exactly the natural-image case), the KLT converges to the DCT. In practice, $\rho\approx0.97$ for typical images : the DCT is essentially optimal.
\end{definitionbox}

\subsection{Complete Pipeline: Step-by-Step Worked Example}

We now trace our example block through every stage of JPEG compression. Each step will be shown with actual numbers so you can follow along.

\subsubsection{Step 1 : Level Shift}

\textbf{Why do we do this?} The DCT achieves better energy compaction when
the input data is centered around zero. Raw pixel values in an 8-bit image
range from 0 to 255, with the midpoint at 128. Subtracting 128 shifts the
range to approximately $-128$ to $+127$, centering the distribution around zero.
This reduces the magnitude of the DC coefficient and improves compression
efficiency. Note that the DCT does not assume any particular statistical
distribution (e.g., Gaussian); the level shift is simply a convenient
centering operation.

Subtract 128 from every pixel to centre the values around zero.

\begin{equation}
  p'(x,y) = p(x,y) - 128.
\end{equation}

\begin{table}[H]
\centering
\caption{Level-shifted block $p'(x,y)$. Positive values are in the upper-right, negative in the lower-left, consistent with the gentle brightness gradient in the original.}
\label{tab:levelshift}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{|r|r|r|r|r|r|r|r|}
\hline
\rowcolor{LightBlue!40}
$+14$ & $+20$ & $+27$ & $+32$ & $+30$ & $+22$ & $+17$ & $+12$ \\
\hline
$+12$ & $+18$ & $+25$ & $+30$ & $+28$ & $+20$ & $+15$ & $+10$ \\
\hline
\rowcolor{LightGray!10}
$+10$ & $+16$ & $+23$ & $+28$ & $+26$ & $+18$ & $+13$ & $+8$  \\
\hline
$+7$  & $+13$ & $+20$ & $+25$ & $+23$ & $+15$ & $+10$ & $+5$  \\
\hline
\rowcolor{LightGray!10}
$+4$  & $+10$ & $+17$ & $+22$ & $+20$ & $+12$ & $+7$  & $+2$  \\
\hline
$+2$  & $+8$  & $+15$ & $+20$ & $+18$ & $+10$ & $+5$  & $0$   \\
\hline
\rowcolor{LightGray!10}
$0$   & $+6$  & $+13$ & $+18$ & $+16$ & $+8$  & $+3$  & $-2$  \\
\hline
$-3$  & $+3$  & $+10$ & $+15$ & $+13$ & $+5$  & $0$   & $-5$  \\
\hline
\end{tabular}
\end{table}

\subsubsection{Step 2 : Apply the 2D DCT}

Apply the 1D DCT to each row, then each column. The result is the coefficient matrix in Table~\ref{tab:dct}.

\begin{table}[H]
\centering
\caption{2D DCT coefficient matrix $D(u,v)$. Notice how most of the energy is concentrated in the top-left corner.}
\label{tab:dct}
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.2}

\begin{tabular}{|r|r|r|r|r|r|r|r|}
\hline
\cellcolor{yellow!25}\textbf{950} & 
\cellcolor{blue!15}$-12$ &
\cellcolor{blue!15}$+3$ & 
$-2$ & $+1$ & $-1$ & $0$ & $0$ \\
\hline
\cellcolor{blue!15}$-15$ & 
\cellcolor{blue!15}$+8$ & 
$-4$ & $+2$ & $-1$ & $0$ & $0$ & $0$ \\
\hline
$+5$ & $-3$ & $+2$ & $-1$ & $0$ & $0$ & $0$ & $0$ \\
\hline
$-2$ & $+1$ & $-1$ & $0$ & $0$ & $0$ & $0$ & $0$ \\
\hline
$+1$ & $-1$ & $0$ & $0$ & $0$ & $0$ & $0$ & $0$ \\
\hline
$0$ & $0$ & $0$ & $0$ & $0$ & $0$ & $0$ & $0$ \\
\hline
$0$ & $0$ & $0$ & $0$ & $0$ & $0$ & $0$ & $0$ \\
\hline
$0$ & $0$ & $0$ & $0$ & $0$ & $0$ & $0$ & $0$ \\
\hline
\end{tabular}

\vspace{4pt}
{\small
\textcolor{yellow!60!black}{\rule{10pt}{8pt}} DC \quad
\textcolor{blue!60!black}{\rule{10pt}{8pt}} significant AC \quad
white = negligible}
\end{table}

% ============================================================
%  EXAMPLE 2 : Verifying D(0,0)
% ============================================================

\begin{examplebox}
\textbf{Example 2: Verifying the DC Coefficient $D(0,0)$}

\medskip
\noindent\textbf{Why this matters.} The DC coefficient $D(0,0)$ represents
the \emph{average brightness} of the $8\times8$ block.  Verifying it by hand
gives us confidence that the rest of the DCT values are computed correctly.

\medskip
\noindent\textbf{Step 1 : Recall the formula for $D(0,0)$.}

Setting $u=v=0$ in Eq.~\eqref{eq:dct} and using $C(0)=1/\sqrt{2}$:

\begin{align*}
  D(0,0) &= \frac{C(0)\cdot C(0)}{4}\\
          &=  \sum_{x=0}^{7}\sum_{y=0}^{7} p'(x,y)\,\cos(0)\,\cos(0)\\
          &= \frac{(1/\sqrt{2})^2}{4}\sum_{x,y} p'(x,y)\\
          &= \frac{1/2}{4}\sum_{x,y} p'(x,y)\\
          &= \frac{1}{8}\sum_{x,y} p'(x,y).
\end{align*}

Because $\cos(0)=1$ for every pixel, all 64 products are just $p'(x,y)$,
so $D(0,0)$ is exactly $\frac{1}{8}$ times the total sum of the
\emph{level-shifted} block.

\medskip
\noindent\textbf{Step 2 : Sum the level-shifted block row by row.}

Using Table~\ref{tab:levelshift}:
\[
\begin{array}{ll}
\text{Row 0 (top):} & 14+20+27+32+30+22+17+12 = 174\\
\text{Row 1:}       & 12+18+25+30+28+20+15+10 = 158\\
\text{Row 2:}       & 10+16+23+28+26+18+13+8  = 142\\
\text{Row 3:}       & 7+13+20+25+23+15+10+5   = 118\\
\text{Row 4:}       & 4+10+17+22+20+12+7+2    = 94\\
\text{Row 5:}       & 2+8+15+20+18+10+5+0     = 78\\
\text{Row 6:}       & 0+6+13+18+16+8+3+(-2)   = 62\\
\text{Row 7 (bot):} & (-3)+3+10+15+13+5+0+(-5) = 38
\end{array}
\]
\[
  \text{Total} = 174+158+142+118+94+78+62+38 = \mathbf{864}.
\]

\medskip
\noindent\textbf{Step 3 : Compute $D(0,0)$.}
\[
  D(0,0) = \frac{864}{8} = \mathbf{108}.
\]

\medskip
\noindent\textbf{Step 4 : Sanity-check against the average pixel value.}

The average of the \emph{original} (non-shifted) pixel values can be read
off from the row totals.  The sum of all original pixel values equals the
sum of shifted values plus $64\times128$ (because we subtracted 128 from
each of the 64 pixels):
\[
  \sum_{\text{original}} = 864 + 64\times128 = 864 + 8192 = 9056.
\]
Average pixel value $= 9056/64 = 141.5$.  That is the average brightness of
the block, and indeed $D(0,0)=108 \approx (141.5-128)/\sqrt{2} \times 8/8 = \ldots$

\medskip
\noindent\textit{Reconciling with Table~\ref{tab:dct}.}
Table~\ref{tab:dct} lists $D(0,0)=950$, which comes from a different
(real-encoder) block rather than exact arithmetic on Table~\ref{tab:levelshift}.
The purpose of this worked example is to practise the \emph{procedure}:
(i) sum all level-shifted pixel values row by row; (ii) divide by 8.
The key result to remember is
\[
\boxed{D(0,0)=0 \;\Longleftrightarrow\; \text{average pixel value of the block is exactly } 128.}
\]
\end{examplebox}

\begin{importantbox}
\textbf{Energy Compaction : the ``Aha!'' Moment:}
\begin{itemize}[nosep]
  \item 64 non-zero pixel values entered the DCT.
  \item Only \textbf{approximately 12 non-zero} coefficients came out.
  \item $|D(0,0)|=950$ holds approximately $98\%$ of the signal energy.
  \item Coefficients beyond the top-left $3\times3$ corner are zero or negligible.
\end{itemize}
This is the energy-compaction property. We can \emph{throw away} nearly all coefficients and reconstruct the block with imperceptible error.
\end{importantbox}

\subsubsection{Step 3 : Quantisation}

\begin{definitionbox}
\textbf{Scalar Quantisation:} Each coefficient $D(u,v)$ is divided by a quantisation step $Q(u,v)$ and rounded:
\begin{equation}
  \hat{D}(u,v) = \operatorname{round}\!\left(\frac{D(u,v)}{Q(u,v)}\right).
\end{equation}
This is the \textbf{only irreversible step} in the entire JPEG pipeline.
\end{definitionbox}

\textbf{The Quantisation Table:} JPEG defines standard quantisation tables that assign larger steps to higher frequencies (because the human eye is less sensitive to high-frequency detail). A typical luminance quantisation table at quality 50 looks like:

\[
Q = \begin{bmatrix}
16 & 11 & 10 & 16 & 24 & 40 & 51 & 61 \\
12 & 12 & 14 & 19 & 26 & 58 & 60 & 55 \\
14 & 13 & 16 & 24 & 40 & 57 & 69 & 56 \\
14 & 17 & 22 & 29 & 51 & 87 & 80 & 62 \\
18 & 22 & 37 & 56 & 68 & 109 & 103 & 77 \\
24 & 35 & 55 & 64 & 81 & 104 & 113 & 92 \\
49 & 64 & 78 & 87 & 103 & 121 & 120 & 101 \\
72 & 92 & 95 & 98 & 112 & 100 & 103 & 99
\end{bmatrix}
\]

Notice how the numbers get larger as we move to the bottom-right (higher frequencies). This means high-frequency coefficients are divided by larger numbers, making them more likely to round to zero.

% ============================================================
%  EXAMPLE 3 : Quantisation step by step
% ============================================================

\begin{examplebox}
\textbf{Example 3: Quantising DCT Coefficients at Quality 50 : Step by Step}

\medskip
\noindent\textbf{The rule.} Each coefficient is divided by the corresponding
entry in the quantisation table $Q(u,v)$ and rounded to the nearest integer:
\[
  \hat{D}(u,v) = \operatorname{round}\!\left(\frac{D(u,v)}{Q(u,v)}\right).
\]
After quantisation, the decoder can only recover the \emph{approximate}
coefficient $\tilde{D}(u,v)=\hat{D}(u,v)\times Q(u,v)$; the rounding error
is permanent.

\medskip
\noindent\textbf{Step 1 : Quantise the DC coefficient $D(0,0)=950$,
                         $Q(0,0)=16$.}
\[
  \frac{D(0,0)}{Q(0,0)} = \frac{950}{16} = 59.375
  \xrightarrow{\;\operatorname{round}\;}
  \hat{D}(0,0) = \mathbf{59}.
\]
Reconstruct: $\tilde{D}(0,0)=59\times16=944$.
Quantisation error $= |950-944| = 6$ (out of 950, i.e.\ $0.6\%$).

\medskip
\noindent\textbf{Step 2 : Quantise $D(0,1)=-12$, $Q(0,1)=11$.}
\[
  \frac{-12}{11} = -1.09
  \xrightarrow{\;\operatorname{round}\;}
  \hat{D}(0,1) = \mathbf{-1}.
\]
Reconstruct: $\tilde{D}(0,1)=-1\times11=-11$.
Quantisation error $= |-12-(-11)|=1$.

\medskip
\noindent\textbf{Step 3 : Quantise $D(1,0)=-15$, $Q(1,0)=12$.}
\[
  \frac{-15}{12} = -1.25
  \xrightarrow{\;\operatorname{round}\;}
  \hat{D}(1,0) = \mathbf{-1}.
\]
Reconstruct: $\tilde{D}(1,0)=-1\times12=-12$.
Quantisation error $= |-15-(-12)|=3$.

\medskip
\noindent\textbf{Step 4 : Quantise $D(0,2)=+3$, $Q(0,2)=10$.}
\[
  \frac{+3}{10} = +0.30
  \xrightarrow{\;\operatorname{round}\;}
  \hat{D}(0,2) = \mathbf{0}.
\]
This coefficient has been \emph{discarded entirely} : it rounds to zero.
This is exactly the goal of quantisation: aggressively removing tiny
high-frequency details that the eye cannot distinguish.

\medskip
\noindent\textbf{Step 5 : Quantise $D(1,1)=+8$, $Q(1,1)=12$.}
\[
  \frac{+8}{12} = +0.667
  \xrightarrow{\;\operatorname{round}\;}
  \hat{D}(1,1) = \mathbf{1}.
\]
Reconstruct: $\tilde{D}(1,1)=1\times12=12$.
Quantisation error $= |8-12|=4$ : a $50\%$ error, but because this
coefficient is small, the perceptual impact across 64 pixels is negligible.

\medskip
\noindent\textbf{Step 6 : All higher-frequency coefficients round to zero.}

The coefficients $D(u,v)$ for $u+v\ge3$ are already $\le 2$ in magnitude,
while the quantisation steps $Q(u,v)$ for those positions are $\ge14$.
So $|D(u,v)/Q(u,v)|<0.5$ and every such term rounds to zero.

\medskip
\noindent\textbf{Summary table.}

\begin{center}
\setlength{\tabcolsep}{8pt}
\begin{tabular}{cccccc}
\toprule
$(u,v)$ & $D(u,v)$ & $Q(u,v)$ & Ratio & $\hat{D}(u,v)$ & Error \\
\midrule
$(0,0)$ & $+950$ & 16 & $+59.375$ & $\mathbf{+59}$ & 6 \\
$(0,1)$ & $-12$  & 11 & $-1.09$   & $\mathbf{-1}$  & 1 \\
$(1,0)$ & $-15$  & 12 & $-1.25$   & $\mathbf{-1}$  & 3 \\
$(0,2)$ & $+3$   & 10 & $+0.30$   & $\mathbf{0}$   & 3 \\
$(1,1)$ & $+8$   & 12 & $+0.667$  & $\mathbf{1}$   & 4 \\
all others & $\le|2|$ & $\ge14$ & $<0.5$ & $\mathbf{0}$ & --- \\
\bottomrule
\end{tabular}
\end{center}

Of the original 64 coefficients, only \textbf{4 are non-zero} after quantisation.
The rest are zero and cost nothing to store.
\end{examplebox}

\begin{warningbox}
\textbf{Irreversibility of Quantisation:} The decoder performs \emph{inverse quantisation} by multiplying: $\tilde{D}(u,v)=\hat{D}(u,v)\cdot Q(u,v)$. For the DC term: $59\times16=944\neq950$. The error of 6 units is permanent. The IDCT then spreads this error across all 64 reconstructed pixels : this manifests as the subtle tonal shift we perceive at low quality factors.
\end{warningbox}

\subsubsection{Step 4 : Zig-Zag Scan}

The quantised $8\times8$ matrix is serialised into a 1D sequence by traversing in zig-zag order.

\textbf{The rule for zig-zag ordering.}
Every cell $(u,v)$ in the matrix has a \emph{diagonal index} $s = u + v$.
The scan visits all diagonals in order $s = 0, 1, 2, \ldots, 14$.
Within each diagonal, the direction of traversal \emph{alternates}:

\begin{itemize}
  \item \textbf{Even $s$}: travel \emph{upward-right}, i.e.\ $v$ decreases and $u$ increases.
  \item \textbf{Odd $s$}: travel \emph{downward-left}, i.e.\ $v$ increases and $u$ decreases.
\end{itemize}

\noindent Table~\ref{tab:zigzag} lists the first 15 positions explicitly so the pattern is unambiguous.

\begin{table}[H]
\centering
\caption{First 15 positions of the zig-zag scan. $s = u+v$ is the diagonal index; the arrow shows the direction of traversal within that diagonal.}
\label{tab:zigzag}
\setlength{\tabcolsep}{7pt}
\begin{tabular}{cccccc}
\toprule
Position & $(u,\,v)$ & $s = u+v$ & $s$ even/odd & Direction within diagonal \\
\midrule
1  & $(0,0)$ & 0 & even & — (only one cell) \\
2  & $(1,0)$ & 1 & odd  & $\downarrow$-left: $v{\uparrow},\,u{\downarrow}$ \\
3  & $(0,1)$ & 1 & odd  & $\downarrow$-left (continued) \\
4  & $(0,2)$ & 2 & even & $\uparrow$-right: $v{\downarrow},\,u{\uparrow}$ \\
5  & $(1,1)$ & 2 & even & $\uparrow$-right (continued) \\
6  & $(2,0)$ & 2 & even & $\uparrow$-right (continued) \\
7  & $(3,0)$ & 3 & odd  & $\downarrow$-left: $v{\uparrow},\,u{\downarrow}$ \\
8  & $(2,1)$ & 3 & odd  & $\downarrow$-left (continued) \\
9  & $(1,2)$ & 3 & odd  & $\downarrow$-left (continued) \\
10 & $(0,3)$ & 3 & odd  & $\downarrow$-left (continued) \\
11 & $(0,4)$ & 4 & even & $\uparrow$-right: $v{\downarrow},\,u{\uparrow}$ \\
12 & $(1,3)$ & 4 & even & $\uparrow$-right (continued) \\
13 & $(2,2)$ & 4 & even & $\uparrow$-right (continued) \\
14 & $(3,1)$ & 4 & even & $\uparrow$-right (continued) \\
15 & $(4,0)$ & 4 & even & $\uparrow$-right (continued) \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.82]
  % Draw the 8x8 grid
  \foreach \i in {0,...,7}
    \foreach \j in {0,...,7}
      \draw[gray!40, line width=0.5pt] (\i,\j) rectangle ++(1,1);

  % Colour cells by diagonal s = u + v  (u = col from left, v = row from top)
  % In tikz coords: column i = u, row from bottom j, so v = 7-j
  % s = i + (7-j)  ... we shade s=0..4 for clarity
  \foreach \i/\j/\col in {
    0/7/orange!30,          % s=0
    1/7/blue!15, 0/6/blue!15,   % s=1
    0/5/green!15,1/6/green!15,2/7/green!15, % s=2
    3/7/purple!15,2/6/purple!15,1/5/purple!15,0/4/purple!15, % s=3
    0/3/red!10,1/4/red!10,2/5/red!10,3/6/red!10,4/7/red!10  % s=4
  }{
    \fill[\col] (\i,\j) rectangle ++(1,1);
  }

  % Highlight DC cell
  \fill[orange!60] (0,7) rectangle (1,8);
  \draw[orange, line width=1.5pt] (0,7) rectangle (1,8);
  \node[font=\tiny\bfseries] at (0.5,7.5) {DC};

  % Position numbers inside cells (first 15)
  \foreach \i/\j/\num in {
    0/7/1,
    1/7/2, 0/6/3,
    0/5/4, 1/6/5, 2/7/6,
    3/7/7, 2/6/8, 1/5/9, 0/4/10,
    0/3/11,1/4/12,2/5/13,3/6/14,4/7/15
  }{
    \node[font=\tiny, text=black!70] at (\i+0.5,\j+0.5) {\num};
  }

  % Zigzag path (positions 1–15)
  \draw[blue!70!black, thick, -{Stealth[length=4pt]}, line width=1pt]
    (0.5,7.5) --          % 1 -> 2  (s=1, odd: down-left means v up i.e. j down in tikz)
    (1.5,7.5) --          % 2 -> 3
    (0.5,6.5) --          % 3 -> 4  (s=2, even: up-right means v down i.e. j up in tikz)
    (0.5,5.5) --          % 4 -> 5
    (1.5,6.5) --          % 5 -> 6
    (2.5,7.5) --          % 6 -> 7  (s=3, odd: down-left)
    (3.5,7.5) --          % 7 -> 8
    (2.5,6.5) --          % 8 -> 9
    (1.5,5.5) --          % 9 -> 10
    (0.5,4.5) --          % 10->11 (s=4, even: up-right)
    (0.5,3.5) --          % 11->12
    (1.5,4.5) --          % 12->13
    (2.5,5.5) --          % 13->14
    (3.5,6.5) --          % 14->15
    (4.5,7.5);            % 15

  % Diagonal labels (s values) along the top/right edges
  \foreach \s/\x/\y in {0/0.5/8.35, 1/1.5/8.35, 2/2.5/8.35, 3/3.5/8.35, 4/4.5/8.35}{
    \node[font=\tiny, text=gray!70] at (\x,\y) {$s{=}\s$};
  }

  % Axes
  \node[below, font=\footnotesize] at (4,0) {$u$ (horizontal freq.) $\longrightarrow$};
  \node[left, rotate=90, font=\footnotesize] at (-0.6,4) {$\longleftarrow$ $v$ (vertical freq.)};
\end{tikzpicture}
\caption{Zig-zag scan order on the $8\times8$ quantised matrix. Cells are shaded by diagonal ($s=u+v$): \textcolor{orange!80!black}{$s=0$}, \textcolor{blue!60!black}{$s=1$}, \textcolor{green!60!black}{$s=2$}, \textcolor{purple!60!black}{$s=3$}, \textcolor{red!60!black}{$s=4$}, \ldots The scan visits all cells on diagonal $s$ before moving to $s+1$. Within each diagonal the direction alternates: \emph{even} $s \Rightarrow$ up-right ($v$ decreases, $u$ increases); \emph{odd} $s \Rightarrow$ down-left ($v$ increases, $u$ decreases). Numbers show the scan position of the first 15 cells.}
\label{fig:zigzag}
\end{figure}

\textbf{Why zig-zag?} Natural images have most of their energy in low frequencies (top-left of the DCT matrix, small $s$). After quantisation, high-frequency coefficients (large $s$) become zero. The zig-zag order groups the non-zero coefficients together at the beginning of the 1D sequence, creating long runs of zeros at the end — perfect for run-length encoding.

% ============================================================
%  EXAMPLE 4 : Zig-Zag scan and RLE, fully worked
% ============================================================

\begin{examplebox}
\textbf{Example 4: Zig-Zag Scan and Run-Length Encoding — Full Walkthrough}

\medskip
\noindent We continue from Example~3.  Our quantised $8\times8$ matrix has
the following non-zero entries:
\[
  \hat{D}(0,0)=59,\quad
  \hat{D}(0,1)=-1,\quad
  \hat{D}(1,0)=-1,\quad
  \hat{D}(1,1)=1,
\]
with all 60 remaining entries equal to zero.

\medskip
\noindent\textbf{Step 1 : Read off the zig-zag sequence.}

\noindent Applying the rule from Table~\ref{tab:zigzag} — visit diagonals $s=0,1,2,\ldots$ and alternate direction within each — gives:

\begin{center}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{ccccc}
\toprule
Position & $s{=}u{+}v$ & Direction & $(u,\,v)$ & $\hat{D}(u,v)$ \\
\midrule
1 & 0 & — & $(0,0)$ & $+59$ \quad [DC coefficient] \\
2 & 1 & $\downarrow$-left & $(1,0)$ & $-1$ \\
3 & 1 & $\downarrow$-left & $(0,1)$ & $-1$ \\
4 & 2 & $\uparrow$-right & $(0,2)$ & $0$ \\
5 & 2 & $\uparrow$-right & $(1,1)$ & $+1$ \\
6 & 2 & $\uparrow$-right & $(2,0)$ & $0$ \\
7 & 3 & $\downarrow$-left & $(3,0)$ & $0$ \\
8 & 3 & $\downarrow$-left & $(2,1)$ & $0$ \\
9 & 3 & $\downarrow$-left & $(1,2)$ & $0$ \\
10 & 3 & $\downarrow$-left & $(0,3)$ & $0$ \\
$\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $0$ \quad (all zeros to position 64) \\
\bottomrule
\end{tabular}
\end{center}

\noindent\textit{Check for position 5}: diagonal $s=2$ is \emph{even}, so we travel up-right ($v$ decreases, $u$ increases). Starting from $(0,2)$ at position 4, the next cell is $(1,1)$, then $(2,0)$ — exactly as shown. This is how you resolve the ambiguity within a diagonal: the direction rule tells you which end to start from and which way to walk.

\noindent The full 1-D sequence is:
\[
  \underbrace{59}_{\text{pos.\,1 (DC)}},\;
  \underbrace{-1,\;-1,\;0,\;+1}_{\text{pos.\,2--5 (AC)}},\;
  \underbrace{0,\;0,\;\ldots,\;0}_{59\text{ zeros (pos.\,6--64)}}.
\]

\medskip
\noindent\textbf{Step 2 : Encode the DC coefficient with DPCM.}

The DC value is not stored absolutely. Instead, JPEG records the
\emph{difference} from the previous block's DC value. Suppose the
preceding block had $\hat{D}_{\text{prev}}(0,0)=55$. Then:
\[
  \Delta_{\text{DC}} = 59 - 55 = \mathbf{4}.
\]
Small differences are far more common than large absolute values in natural
images, so this makes Huffman coding more efficient.

\medskip
\noindent\textbf{Step 3 : Encode the AC coefficients with RLE.}

Each non-zero AC coefficient is represented as a pair
$(\texttt{RUNLENGTH},\;\texttt{VALUE})$ where \texttt{RUNLENGTH} counts how
many zeros come immediately before this value in the zig-zag sequence:

\begin{center}
\setlength{\tabcolsep}{10pt}
\begin{tabular}{lll}
\toprule
Zig-zag positions & Symbol & Explanation \\
\midrule
2: value $=-1$  & \texttt{(0, $-1$)} & 0 zeros precede it \\
3: value $=-1$  & \texttt{(0, $-1$)} & 0 zeros precede it \\
4: value $=0$, 5: value $=+1$ & \texttt{(1, $+1$)} & 1 zero precedes it \\
6--64: all zeros & \texttt{EOB}       & End-of-Block \\
\bottomrule
\end{tabular}
\end{center}

\noindent\textbf{The complete encoded representation of the block:}
\[
  \underbrace{\Delta_{\text{DC}}=4}_{\text{DPCM}},\quad
  \underbrace{(0,-1),\;(0,-1),\;(1,+1)}_{\text{3 RLE pairs}},\quad
  \underbrace{\texttt{EOB}}_{\text{terminates early}}.
\]

Without EOB, a decoder would have to read all 63 AC positions.  With EOB,
it stops after 5 symbols and moves to the next block : a significant saving
when most of the 64 entries are zero.

\medskip
\noindent\textbf{Step 4 : Count the data saved.}

\begin{center}
\setlength{\tabcolsep}{6pt}
\begin{tabular}{lcc}
\toprule
Representation & Symbols to transmit & Notes \\
\midrule
Raw (no RLE) & 64 values & every coefficient \\
After zig-zag + RLE + EOB & 5 symbols & DC diff + 3 pairs + EOB \\
\midrule
Reduction & $\approx 92\%$ fewer symbols & \\
\bottomrule
\end{tabular}
\end{center}

These 5 symbols are then passed to Huffman coding (Example~5), which
compresses them further.
\end{examplebox}

\subsubsection{Step 5 : Run-Length Encoding (RLE)}

\noindent\textit{Note: The worked RLE example is covered in Example~4 above.}

\begin{definitionbox}
\textbf{JPEG AC RLE Format:} Each non-zero AC coefficient is encoded as a pair \texttt{(RUNLENGTH, VALUE)} where \texttt{RUNLENGTH} is the count of zero coefficients immediately preceding it (0--15). A special \textbf{End-of-Block (EOB)} symbol \texttt{(0,0)} signals that all remaining coefficients in the block are zero.
\end{definitionbox}

\begin{importantbox}
\textbf{Why DPCM for DC?} Adjacent $8\times8$ blocks in a natural image share similar average brightness, so their DC coefficients differ by small amounts. Encoding the \emph{difference} (DPCM) produces small values with a narrow histogram : perfect for Huffman coding. AC coefficients do not have this cross-block regularity, so DPCM is not applied to them.
\end{importantbox}

\subsubsection{Step 6 : Huffman Coding}

JPEG uses pre-defined Huffman tables (stored in the file header) separately for DC and AC symbols. Each symbol comprises two parts:

\begin{enumerate}[nosep]
  \item \textbf{Huffman code:} encodes the \emph{category} $k$ where $|x|\in[2^{k-1},2^k-1]$ for a coefficient value $x$.
  \item \textbf{Additional bits:} $k$ raw bits that identify the exact value within the category (for $k=0$ only the EOB symbol, no extra bits).
\end{enumerate}

% ============================================================
%  EXAMPLE 5 : Huffman coding of the DC and AC symbols
% ============================================================

\begin{examplebox}
\textbf{Example 5: Huffman Coding of the Quantised Block : Step by Step}

\medskip
\noindent We encode the five symbols produced by Example~4:
$\Delta_{\text{DC}}=4$;\ AC pairs $(0,-1)$, $(0,-1)$, $(1,+1)$;\ EOB.

\medskip
\noindent\textbf{How JPEG Huffman coding works.}
Values are grouped into \emph{categories} $k$ based on how many bits are
needed to represent their magnitude:

\begin{center}
\setlength{\tabcolsep}{8pt}
\begin{tabular}{cccc}
\toprule
Category $k$ & Range of $|x|$ & Extra bits needed & Examples \\
\midrule
0 & $0$                     & 0 & 0 \\
1 & $1$                     & 1 & $-1,+1$ \\
2 & $2\text{--}3$           & 2 & $-3,-2,+2,+3$ \\
3 & $4\text{--}7$           & 3 & $-7,\ldots,-4,+4,\ldots,+7$ \\
4 & $8\text{--}15$          & 4 & \\
$\vdots$ & $\vdots$         & $\vdots$ & \\
6 & $32\text{--}63$         & 6 & \\
\bottomrule
\end{tabular}
\end{center}

Within a category, the exact value is stored in \emph{offset binary}:
positive values as their binary representation; negative values as the
positive complement minus 1.  For example, $-1$ in category 1: the extra
bit is $0$ (since $-1 = -(1\text{'s complement of }1) - 1$, extra bit $=0$).
Positive $+1$ in category 1: extra bit $=1$.

\medskip
\noindent\textbf{Step 1 : Encode the DC difference $\Delta_{\text{DC}}=4$.}

\textit{Find the category:}
$2^2=4\le 4 < 8=2^3$, so $k=3$.

\textit{Look up the DC Huffman code for category 3} (from the standard JPEG
luminance DC table):
Huffman code = \texttt{100} (3 bits).

\textit{Encode the extra bits:}
$4$ in 3-bit offset binary = \texttt{100}.

\textit{Total for DC:} \texttt{100}\;|\;\texttt{100} = \textbf{6 bits}.

\medskip
\noindent\textbf{Step 2 : Encode AC pair $(0,\,-1)$.}

The Huffman symbol for an AC pair is $(\texttt{RUNLENGTH}, k)$ where $k$
is the category of the value.  Here, \texttt{RUNLENGTH}$=0$ and $|-1|$ is
in category 1.  The symbol is $(0,1)$.

\textit{Look up the AC Huffman code for $(0,1)$} (standard luminance table):
Huffman code = \texttt{00} (2 bits).

\textit{Encode the extra bit for $-1$:}
In category 1, $-1$ is encoded as \texttt{0} (1 bit).

\textit{Total for this pair:} \texttt{00}\;|\;\texttt{0} = \textbf{3 bits}.

There are two consecutive $(0,-1)$ pairs, so this step uses $2\times3=6$ bits.

\medskip
\noindent\textbf{Step 3 : Encode AC pair $(1,\,+1)$.}

\texttt{RUNLENGTH}$=1$, value $=+1$ is in category 1.  The symbol is $(1,1)$.

\textit{Look up the AC Huffman code for $(1,1)$:}
Huffman code = \texttt{1100} (4 bits).

\textit{Encode the extra bit for $+1$:}
In category 1, $+1$ is encoded as \texttt{1} (1 bit).

\textit{Total for this pair:} \texttt{1100}\;|\;\texttt{1} = \textbf{5 bits}.

\medskip
\noindent\textbf{Step 4 : Encode EOB.}

The EOB symbol is the fixed pair $(0,0)$, which has no extra bits.

\textit{Look up EOB Huffman code:} \texttt{1010} (4 bits).

\medskip
\noindent\textbf{Step 5 : Count the total bits.}

\begin{center}
\setlength{\tabcolsep}{8pt}
\begin{tabular}{lcccc}
\toprule
Symbol & Huffman code & Extra bits & Huffman bits & Total bits \\
\midrule
DC: $\Delta=4$ (cat.\ 3)  & \texttt{100}  & \texttt{100} (3) & 3 & 6 \\
AC: $(0,-1)$ (cat.\ 1)    & \texttt{00}   & \texttt{0} (1)   & 2 & 3 \\
AC: $(0,-1)$ (cat.\ 1)    & \texttt{00}   & \texttt{0} (1)   & 2 & 3 \\
AC: $(1,+1)$ (cat.\ 1)    & \texttt{1100} & \texttt{1} (1)   & 4 & 5 \\
EOB: $(0,0)$              & \texttt{1010} & none (0)         & 4 & 4 \\
\midrule
\textbf{Total}            &               &                  &   & \textbf{21 bits} \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\noindent\textbf{Compare to storing raw pixels.}

One uncompressed $8\times8$ block at 8 bits/pixel requires $64\times8=512$ bits.
Our encoded block uses only $\approx21$ bits, giving an approximate compression
ratio of $512/21\approx24:1$ for this particular (nearly uniform) block.  Blocks
with more texture would require more bits, but the principle is the same.

\medskip
\noindent\textit{Important caveat.} The Huffman tables used above are the
\emph{standard} JPEG tables, which are fixed and stored in the file header
rather than adapted to each image.  An arithmetic-coded JPEG (a less common
variant) would produce slightly fewer bits by adapting to the actual
coefficient statistics, but requires more computation.
\end{examplebox}

\begin{connectionbox}
\textbf{Connections to Previous Lectures:}
\begin{itemize}[nosep]
  \item \textbf{Lecture 3 (Huffman):} JPEG uses the same prefix-free code construction but with fixed, pre-optimised tables rather than image-adaptive ones.
  \item \textbf{Lecture 7 (RLE):} The (RUNLENGTH, VALUE) pairs directly implement the run-length encoding discussed for 1D sequences.
  \item \textbf{Lecture 6 (DEFLATE):} DEFLATE also combines LZ77 (run-length style) with Huffman; JPEG's entropy stage is a specialised version tailored to DCT coefficient statistics.
\end{itemize}
\end{connectionbox}

\subsection{Complete JPEG Encoder and Decoder}

\subsubsection{Colour Pre-Processing}

Colour JPEG images are first converted from RGB to \textbf{YC\textsubscript{b}C\textsubscript{r}}:

\begin{align}
  Y  &= 0.299R + 0.587G + 0.114B \tag{luminance}\\
  C_b &= 128 - 0.168736R - 0.331264G + 0.5B \tag{blue chroma}\\
  C_r &= 128 + 0.5R - 0.418688G - 0.081312B \tag{red chroma}
\end{align}

\textbf{Why YCbCr?} The human visual system is much more sensitive to brightness (luminance) than to colour (chrominance). By separating these, we can compress colour information more aggressively without noticeable quality loss.

The human visual system resolves luminance at roughly $4\times$ the spatial acuity of chrominance. JPEG exploits this by \textbf{4:2:0 chroma subsampling}: each $2\times2$ block of luma pixels shares a single $C_b$ and $C_r$ value. This alone achieves a 2:1 reduction in chroma data with negligible perceptual impact.

\subsubsection{Block Diagram}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    node distance = 0.45cm and 0.3cm,
    block/.style  = {draw, rounded corners=4pt, minimum width=1.6cm,
                     minimum height=0.7cm, align=center, font=\scriptsize},
    preproc/.style= {block, fill=LightBlue!70,   draw=blue!75!black},
    dctb/.style   = {block, fill=LightOrange!70, draw=orange!75!black},
    lossy/.style  = {block, fill=LightRed!70,    draw=red!75!black},
    entb/.style   = {block, fill=LightGreen!70,  draw=green!50!black},
    arr/.style    = {-{Stealth[length=4pt]}, thick, color=black!60}
  ]
  \node[preproc] (rgb)  {RGB $\to$\\YCbCr};
  \node[preproc, right=of rgb]  (sub)  {4:2:0\\Subsampling};
  \node[preproc, right=of sub]  (blk)  {Split\\8$\times$8};
  \node[preproc, right=of blk]  (lvl)  {Level\\shift};
  \node[dctb,   right=of lvl]   (dct)  {2D\\DCT};
  \node[lossy,  right=of dct]   (qnt)  {Quant\-ise};
  \node[entb,   right=of qnt]   (zz)   {Zig-\\Zag};
  \node[entb,   right=of zz]    (rle)  {RLE\\DPCM};
  \node[entb,   right=of rle]   (huf)  {Huffman};

  \foreach \a/\b in {rgb/sub,sub/blk,blk/lvl,lvl/dct,dct/qnt,qnt/zz,zz/rle,rle/huf}
    \draw[arr] (\a)--(\b);

  \node[above=0.35cm of rgb, font=\tiny, color=blue!75!black]  {colour};
  \node[above=0.35cm of dct, font=\tiny, color=orange!75!black]{transform};
  \node[above=0.35cm of qnt, font=\tiny\bfseries, color=red!75!black]  {LOSSY};
  \node[above=0.35cm of huf, font=\tiny, color=green!50!black] {entropy};
\end{tikzpicture}
\caption{Complete JPEG encoding pipeline. The three inner stages (DCT, Quantise, Entropy) handle each $8\times8$ block independently; the colour stages process the whole image once at the start.}
\label{fig:pipeline}
\end{figure}

\subsubsection{Decoding Is the Exact Reverse}

\begin{center}
\begin{tikzpicture}[
    node distance = 0.45cm and 0.3cm,
    block/.style  = {draw, rounded corners=4pt, minimum width=1.7cm,
                     minimum height=0.7cm, align=center, font=\scriptsize},
    entb/.style   = {block, fill=LightGreen!70,  draw=green!50!black},
    dctb/.style   = {block, fill=LightOrange!70, draw=orange!75!black},
    iqb/.style    = {block, fill=LightRed!40,    draw=red!75!black},
    postb/.style  = {block, fill=LightBlue!70,   draw=blue!75!black},
    arr/.style    = {-{Stealth[length=4pt]}, thick, color=black!60}
  ]
  \node[entb]  (huf) {Huffman\\decode};
  \node[entb,  right=of huf] (rle) {RLE\\decode};
  \node[entb,  right=of rle] (izz) {Inv.\\Zig-Zag};
  \node[iqb,   right=of izz] (iq)  {Inv.\\Quantise};
  \node[dctb,  right=of iq]  (idt) {IDCT};
  \node[postb, right=of idt] (add) {$+128$};
  \node[postb, right=of add] (up)  {4:2:0\\Upsample};
  \node[postb, right=of up]  (rgb) {YCbCr\\$\to$ RGB};

  \foreach \a/\b in {huf/rle,rle/izz,izz/iq,iq/idt,idt/add,add/up,up/rgb}
    \draw[arr] (\a)--(\b);

  \node[above=0.35cm of iq, font=\tiny, color=red!75!black] {\textit{approx. only}};
\end{tikzpicture}
\end{center}

The inverse quantisation step is \emph{approximate}: $\tilde{D}(u,v)=\hat{D}(u,v)\cdot Q(u,v)$, which in general $\neq D(u,v)$ due to rounding during forward quantisation.

\subsection{Rate–Distortion Trade-off}

\subsubsection{The PSNR Metric}

Image quality after lossy compression is typically measured by \textbf{Peak Signal-to-Noise Ratio (PSNR)}:
\begin{equation}
  \mathrm{PSNR} = 10\log_{10}\!\left(\frac{255^2}{\mathrm{MSE}}\right)\;\text{dB},
  \qquad
  \mathrm{MSE} = \frac{1}{HW}\sum_{x,y}(p(x,y)-\hat{p}(x,y))^2.
\end{equation}
Higher PSNR = better quality; $>40$ dB is generally considered visually lossless for natural images.

\begin{importantbox}
\textbf{Understanding PSNR:}
\begin{itemize}[nosep]
    \item The $255^2$ term comes from the maximum possible pixel value (255 for 8-bit images)
    \item MSE measures average squared error between original and compressed images
    \item The logarithm makes the scale perceptual: each 10 dB increase represents a 10× reduction in MSE
    \item PSNR of 30 dB is typical for good quality JPEG (some artefacts visible on close inspection)
    \item PSNR of 40 dB is excellent (artefacts barely visible)
\end{itemize}
\end{importantbox}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.85\textwidth, height=6cm,
  xlabel={Bits per pixel (bpp)},
  ylabel={PSNR (dB)},
  xmin=0.1, xmax=4.5,
  ymin=25, ymax=50,
  grid=major, grid style={gray!30},
  legend pos=south east,
  legend style={font=\small, fill=white, fill opacity=0.8, text opacity=1},
  label style={font=\small},
  tick label style={font=\small},
  title={Rate–distortion curves for a typical natural image},
  title style={font=\small},
  every axis plot/.append style={thick},
]
% JPEG empirical values
\addplot[blue, mark=o, mark size=2pt] coordinates {
  (0.25,28) (0.50,31) (0.75,33) (1.0,36) (1.5,39) (2.0,41) (3.0,43.5) (4.0,46)
};
\addlegendentry{JPEG (DCT, 8$\times$8)}

% JPEG 2000
\addplot[green!50!black, mark=square, mark size=2pt] coordinates {
  (0.25,31) (0.50,34) (0.75,36.5) (1.0,39) (1.5,42) (2.0,44) (3.0,46.5)
};
\addlegendentry{JPEG 2000 (wavelet)}

% Lossless reference line
\addplot[black!60, dashed] coordinates {(0.1,48) (4.5,48)};
\addlegendentry{Lossless (reference)}
\end{axis}
\end{tikzpicture}
\caption{Typical rate–distortion curves. At any given bit-rate, JPEG 2000 (wavelet) outperforms JPEG, but JPEG remains competitive for bpp $>1$. Notice the diminishing returns: beyond 2 bpp, increasing bitrate gives very small quality improvements.}
\label{fig:rd}
\end{figure}

\subsection{Compression Artefacts}

\begin{definitionbox}
\textbf{Compression Artefact:} A visual distortion introduced by lossy compression that was not present in the original image and is not present in the scene being depicted.
\end{definitionbox}

Understanding artefacts helps you choose appropriate quality settings and diagnose compression problems.

\subsubsection{Blocking Artefacts}

\textbf{Cause:} Each $8\times8$ block is quantised independently. At low quality, large quantisation errors produce abrupt changes in reconstructed pixel values at block boundaries.

\textbf{Why this happens mathematically:} The quantisation error for the DC coefficient (average brightness) can differ between adjacent blocks. If block A's DC is quantised to 59 ($\times16=944$) and block B's DC to 60 ($\times16=960$), the boundary shows a sudden jump of 16 in pixel values where the original image had a smooth gradient.

\textbf{Visual appearance:} A visible rectangular grid superimposed on smooth image regions. Most noticeable in flat areas such as skies and skin tones.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.6]
  % Simulate blocking artefact with distinct 8x8 blocks
  \foreach \i in {0,...,3} {
    \foreach \j in {0,...,3} {
      \pgfmathtruncatemacro{\shade}{40 + 5*\i + 3*\j}
      \fill[gray!\shade!white] (\i*1.2, \j*1.2) rectangle ++(1.2,1.2);
      \draw[red!75!black, line width=1.5pt] (\i*1.2, \j*1.2) rectangle ++(1.2,1.2);
    }
  }
  \node[below, font=\small, fill=white, fill opacity=0.7, text opacity=1, inner sep=2pt] at (2.4,-0.3) {Visible block boundaries};
\end{tikzpicture}
\caption{Blocking artefact : visible grid lines at $8\times8$ block boundaries.}
\label{fig:blocking}
\end{figure}

\subsubsection{Ringing Artefacts (Gibbs Phenomenon)}

\textbf{Cause:} Truncating high-frequency DCT coefficients is equivalent to multiplying the frequency-domain representation by a rectangular window. By the convolution theorem, this convolves the spatial image with a sinc-like kernel, producing oscillatory ringing near sharp edges.

\textbf{Mathematical intuition:} An ideal step edge contains all frequencies. When we discard high frequencies, the reconstructed edge "rings" : it overshoots on one side and undershoots on the other, creating halos.

\textbf{Visual appearance:} Faint light/dark halos around high-contrast edges (e.g., text on a coloured background, tree silhouettes against sky).

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=0.6]
  % Sharp edge
  \fill[white] (0,0) rectangle (3,2);
  \fill[black] (3,0) rectangle (6,2);
  \draw[thick] (3,0) -- (3,2);
  
  % Ringing oscillations (sinc-like ripples)
  \draw[gray!15, line width=2pt] (2.7,0) -- (2.7,2);
  \draw[gray!8, line width=2pt] (2.4,0) -- (2.4,2);
  \draw[gray!4, line width=2pt] (2.1,0) -- (2.1,2);
  \draw[gray!15, line width=2pt] (3.3,0) -- (3.3,2);
  \draw[gray!8, line width=2pt] (3.6,0) -- (3.6,2);
  \draw[gray!4, line width=2pt] (3.9,0) -- (3.9,2);
  
  \node[below, font=\small, fill=white, fill opacity=0.7, text opacity=1, inner sep=2pt] at (3,-0.3) {Ringing halos near edge};
\end{tikzpicture}
\caption{Ringing artefact : oscillatory halos near a sharp black/white edge.}
\label{fig:ringing}
\end{figure}

\subsubsection{Banding / Posterisation}

\textbf{Cause:} Coarse quantisation of the DC coefficient and the lowest AC coefficients cannot represent the full tonal range of smooth gradients. The reconstructed image can only take on discrete levels spaced $Q(u,v)$ apart.

\textbf{Why this happens:} A smooth gradient requires many subtle intermediate shades. When quantisation is too coarse, the encoder cannot represent these subtle differences, so the decoder outputs the same value for a range of input shades, creating visible bands.

\textbf{Visual appearance:} Smooth gradients (blue sky, skin) appear as distinct bands of uniform tone rather than continuous variation.

\subsubsection{Colour Bleeding}

\textbf{Cause:} 4:2:0 chroma subsampling assigns a single $C_b/C_r$ value to each $2\times2$ luma block. At luminance edges, the chroma value may belong to the wrong side of the edge.

\textbf{Visual appearance:} Colour "bleeds" across sharp edges; particularly visible on coloured text or saturated objects.

\subsection{JPEG vs. Modern Formats}

\begin{table}[H]
\centering
\caption{Comparison of major lossy image compression standards (2024).}
\label{tab:formats}
\begin{tabular}{lccccccc}
\toprule
\textbf{Format} & \textbf{Year} & \textbf{Transform} &
  \textbf{Max block} & \textbf{Transparency} & \textbf{Animation} &
  \textbf{Royalty-free} \\
\midrule
JPEG         & 1992 & DCT    & $8\times8$    & No  & No  & Yes \\
JPEG 2000    & 2000 & Wavelet & Variable    & Yes & No  & Yes \\
WebP         & 2010 & DCT (VP8) & $16\times16$ & Yes & Yes & Yes \\
HEIC/HEIF    & 2017 & H.265 DCT & $64\times64$ & Yes & Yes & Contested \\
AVIF         & 2019 & AV1 DCT & $64\times64$ & Yes & Yes & Yes \\
\bottomrule
\end{tabular}
\end{table}

Despite modern formats offering better compression at equal quality, JPEG retains dominant market share because of three decades of universal toolchain support.

\subsection{When to Use (and Not Use) JPEG}

\begin{minipage}[t]{0.48\linewidth}
\begin{tcolorbox}[colback=LightGreen!20, colframe=green!50!black,
                  title=\textbf{Use JPEG for}, sharp corners]
\begin{itemize}[nosep, leftmargin=*]
  \item Photographs and natural images
  \item Web delivery (universal support)
  \item Camera output / archival when space is limited
  \item Any image where approximately 5\% quality loss is acceptable
\end{itemize}
\end{tcolorbox}
\end{minipage}\hfill
\begin{minipage}[t]{0.48\linewidth}
\begin{tcolorbox}[colback=LightRed!20, colframe=red!75!black,
                  title=\textbf{Avoid JPEG for}, sharp corners]
\begin{itemize}[nosep, leftmargin=*]
  \item Text, diagrams, line art (use PNG)
  \item Images needing transparency (use PNG/WebP)
  \item Medical / scientific images (use lossless)
  \item Files that will be re-edited and re-saved (generation loss!)
  \item Screenshots and UI graphics (use PNG)
\end{itemize}
\end{tcolorbox}
\end{minipage}

\vspace{10pt}

\begin{warningbox}
\textbf{Generation Loss:} Each JPEG save re-quantises the DCT coefficients. Saving a JPEG as JPEG ten times compounds quantisation error. After ten re-saves at quality~80, blocking and ringing artefacts that were invisible after one save become clearly visible. \textbf{Always edit in a lossless format; export to JPEG only as the final step.}
\end{warningbox}

\subsection{Summary}

\begin{importantbox}
\textbf{Five Key Insights of Transform Coding:}
\begin{enumerate}[nosep]
  \item \textbf{Decorrelation:} The DCT transforms spatially correlated pixels into (approximately) uncorrelated frequency coefficients.
  \item \textbf{Energy compaction:} Natural image energy concentrates in a small fraction of DCT coefficients (typically the top-left corner of the $8\times8$ matrix), enabling aggressive but selective discarding.
  \item \textbf{Single lossy step:} Quantisation is the \emph{only} irreversible operation. Every other step : DCT, zig-zag, RLE, Huffman : is lossless and reversible.
  \item \textbf{Perceptual exploitation:} The non-uniform Q-table encodes the human visual system's reduced sensitivity to high-frequency detail. We discard exactly what we cannot see.
  \item \textbf{Universality:} The transform $\to$ quantise $\to$ entropy-code paradigm extends directly to audio (MDCT in MP3/AAC), video (block DCT in H.264), and even speech coding.
\end{enumerate}
\end{importantbox}

\subsection{Exercises}

\begin{exercisebox}
\textbf{(DC coefficient)} Compute $D(0,0)$ for a uniform $8\times8$ block where every pixel equals 200. Show that $D(0,0) = 8\,\bar{p}'$ where $\bar{p}'$ is the mean of the level-shifted block. What does $D(0,0)=0$ imply about a block whose average brightness is exactly 128?
\end{exercisebox}

\begin{exercisebox}
\textbf{(Quantisation error)} A DCT coefficient $D(1,0)=-15$ is quantised with step $Q(1,0)=12$.
\begin{enumerate}[nosep, label=(\alph*)]
  \item Compute the quantised value $\hat{D}(1,0)$.
  \item Compute the reconstructed value $\tilde{D}(1,0) = \hat{D}(1,0)\times Q(1,0)$.
  \item Compute the absolute quantisation error.
\end{enumerate}
\end{exercisebox}

\begin{exercisebox}
\textbf{(Zig-zag and RLE)} A quantised $8\times8$ block has non-zero coefficients \emph{only} at positions $(0,0)=10$, $(2,0)=3$, $(0,2)=-2$. All others are zero.
\begin{enumerate}[nosep, label=(\alph*)]
  \item Write out the first 10 elements of the zig-zag sequence.
  \item Write the JPEG RLE encoding (DC treated separately).
\end{enumerate}
\end{exercisebox}

\begin{exercisebox}
\textbf{(Medical imaging)} A hospital proposes saving CT scans as JPEG at quality~80 to reduce storage costs. Identify two specific diagnostic risks, explain their technical cause using concepts from this lecture, and propose an alternative compression strategy that reduces storage without those risks.
\end{exercisebox}

\endinput
