\documentclass[11pt,a4paper]{article}
\input{lecture-preamble}

\title{\vspace{-1.3cm}\textbf{Lecture 1: Mathematical Foundations\\ for Lattice-Based Cryptography}\\[4pt]
\large Groups, Rings, Fields, Modular Arithmetic,\\ Vector Spaces, Basis, and Lattices}
\author{Dr.\ Faisal Aslam}
\date{Lecture Notes --- Session 1 (approx. 2 hours)}

\begin{document}
\maketitle
\vspace{-0.8cm}

\begin{notebox}
\small\textbf{Purpose of this lecture.} Every technical term used later in the thesis proposal --- ``lattice,'' ``modulus,'' ``ring $\mathbb{Z}_q[x]/(x^n+1)$'' --- rests on a small set of algebraic building blocks. This lecture builds those blocks, one at a time, from the ground up: \emph{Group $\to$ Subgroup $\to$ Ring $\to$ Field $\to$ Modular Arithmetic $\to$ Vector Space $\to$ Basis $\to$ Lattice}. Nothing here requires prior abstract algebra; every definition is followed immediately by concrete, worked examples.
\end{notebox}

\tableofcontents
\vspace{0.3cm}

\section{Why Start With Algebra?}

Lattice-based cryptography is, at its core, about doing arithmetic on \emph{vectors} whose coordinates live in a \emph{finite, wraparound number system}, arranged according to strict algebraic rules. Before we can even state what a lattice is, we need precise answers to questions like:
\begin{itemize}[leftmargin=*]
  \item What does it mean for a set of numbers to have a well-behaved notion of ``add'' and ``multiply''? (\textbf{Groups, Rings, Fields})
  \item What happens when arithmetic ``wraps around,'' like a clock? (\textbf{Modulus})
  \item What does it mean to build every point in a space out of a small number of building-block directions? (\textbf{Vector spaces, Basis})
  \item What happens when we only allow \emph{whole-number} combinations of those building blocks, instead of any real-number combination? (\textbf{Lattice})
\end{itemize}
This lecture answers each of these in order. By the end, the definition of a lattice (Section~8) should feel like the natural, inevitable next step, not an isolated new idea.

\section{Sets and Binary Operations (Quick Refresher)}

A \textbf{set} is just a collection of objects (numbers, vectors, symbols --- anything). A \textbf{binary operation} on a set $S$ is a rule that takes two elements of $S$ and combines them into a third element, e.g., ordinary addition $+$ takes two numbers and produces their sum. We write $a * b$ for a general binary operation.

An operation is \textbf{closed} on $S$ if combining two elements of $S$ always gives another element of $S$ (never something outside $S$). This sounds obvious but fails more often than you'd expect --- see Example~1 below.

\begin{exbox}{Closure Can Fail}
Let $S = \{1, 2, 3, 4, 5\}$ (just these five numbers) and let $*$ be ordinary addition. Then $2 * 4 = 6$, but $6 \notin S$. So ordinary addition is \emph{not} closed on $S$. This is exactly the kind of problem modular arithmetic (Section~5) is built to fix.
\end{exbox}

\section{Groups}

\begin{defbox}{Group}
A set $G$ together with a binary operation $*$ is called a \textbf{group}, written $(G, *)$, if all four of the following hold:
\begin{enumerate}[leftmargin=*, label=(G\arabic*)]
  \item \textbf{Closure:} for all $a, b \in G$, $a * b \in G$.
  \item \textbf{Associativity:} for all $a,b,c \in G$, $(a*b)*c = a*(b*c)$.
  \item \textbf{Identity:} there exists an element $e \in G$ such that $a * e = e * a = a$ for all $a \in G$.
  \item \textbf{Inverses:} for every $a \in G$, there exists $a^{-1} \in G$ such that $a * a^{-1} = a^{-1} * a = e$.
\end{enumerate}
If, in addition, $a * b = b * a$ for all $a, b \in G$ (the order doesn't matter), the group is called \textbf{abelian} (commutative).
\end{defbox}

\begin{exbox}{Groups You Already Know}
\begin{itemize}[leftmargin=*]
  \item $(\mathbb{Z}, +)$ --- the integers under addition. Identity is $0$; the inverse of $a$ is $-a$. This is an abelian group.
  \item $(\mathbb{R} \setminus \{0\}, \times)$ --- nonzero real numbers under multiplication. Identity is $1$; the inverse of $a$ is $1/a$. Also abelian. (We had to remove $0$ because $0$ has no multiplicative inverse.)
  \item $(\mathbb{Z}, \times)$ --- integers under multiplication --- is \textbf{not} a group: $2$ has no inverse in $\mathbb{Z}$ (we'd need $1/2$, which isn't an integer). This fails axiom (G4).
\end{itemize}
\end{exbox}

\begin{rmkbox}{Why Cryptography Cares About Groups}
Many classical cryptographic schemes (Diffie--Hellman, ElGamal, elliptic-curve cryptography) are built directly on top of group structure --- the security relies on certain problems (like ``given $g$ and $g^a$, find $a$'') being hard inside a specific group. Lattice-based cryptography instead builds on richer structures (rings and modules, coming next), but the group is always the first layer underneath.
\end{rmkbox}

\begin{exercisebox}{Try This Before Next Lecture}
Is $(\mathbb{Z}_5 \setminus \{0\}, \times \bmod 5) = (\{1,2,3,4\}, \times \bmod 5)$ a group? Check all four axioms directly by building the multiplication table. (You'll need Section~6 for what ``$\bmod 5$'' means if it's new to you --- come back to this after reading that section.)
\end{exercisebox}

\subsection{Subgroups}

There is one more group-related idea we need before we can properly define a lattice: a group living \emph{inside} another group.

\begin{defbox}{Subgroup}
Let $(G, *)$ be a group. A subset $H \subseteq G$ is a \textbf{subgroup} of $G$ if $H$ is itself a group under the same operation $*$. In practice, this is easiest to check with a simple test: $H$ is a subgroup of $G$ if and only if (1) $H$ is nonempty (usually: it contains the identity $e$), (2) $H$ is closed under $*$ (for $a,b \in H$, $a*b \in H$), and (3) $H$ is closed under inverses (for $a \in H$, $a^{-1} \in H$). Associativity is automatic, since it already held for all of $G$.
\end{defbox}

\begin{exbox}{Subgroups You Can Picture}
\begin{itemize}[leftmargin=*]
  \item For any integer $n$, the set $n\mathbb{Z} = \{\ldots, -2n, -n, 0, n, 2n, \ldots\}$ (all multiples of $n$) is a subgroup of $(\mathbb{Z}, +)$: it contains $0$, adding two multiples of $n$ gives another multiple of $n$, and $-(kn)$ is also a multiple of $n$. For example, $2\mathbb{Z}$ is ``all even integers.''
  \item The natural numbers $\mathbb{N} = \{0,1,2,\ldots\}$ are \textbf{not} a subgroup of $(\mathbb{Z},+)$: $\mathbb{N}$ is closed under addition, but it fails closure under inverses --- $1 \in \mathbb{N}$, but $-1 \notin \mathbb{N}$.
\end{itemize}
\end{exbox}

\begin{previewbox}{Why This Matters in Two Sections}
Keep the example $n\mathbb{Z} \subseteq \mathbb{Z}$ in mind. In Section~8, we will define a lattice as a special kind of subgroup of $\mathbb{R}^n$ --- and $n\mathbb{Z}$ is exactly a simple, one-dimensional lattice in disguise: an evenly-spaced, discrete set of points, closed under addition and negation, sitting inside the larger group $\mathbb{Z}$ (or $\mathbb{R}$).
\end{previewbox}

\subsection{Cosets and Quotient Groups: Where $\mathbb{Z}_n$ Really Comes From}

Section~6 will introduce $\mathbb{Z}_n$ informally, as ``remainders with wraparound arithmetic.'' That's the right intuition, but it's worth seeing the fully rigorous construction now, because it is a direct, satisfying payoff of the subgroup idea --- and it is the same construction (``take a group, collapse it by a subgroup'') that shows up again later when we build the ring $R_q = \mathbb{Z}_q[x]/(x^n+1)$ that Kyber and Dilithium actually use.

\begin{defbox}{Coset}
Let $H$ be a subgroup of an abelian group $(G,+)$. For $a \in G$, the \textbf{coset} $a + H$ is the set $\{a + h : h \in H\}$ --- ``shift $H$ by $a$.'' Two elements $a,b \in G$ turn out to produce the \emph{same} coset, $a+H = b+H$, exactly when $a - b \in H$. This gives a clean way to sort all of $G$ into non-overlapping ``buckets,'' each one a shifted copy of $H$.
\end{defbox}

\begin{exbox}{The Cosets of $3\mathbb{Z}$ in $\mathbb{Z}$}
Take $G = \mathbb{Z}$ and $H = 3\mathbb{Z} = \{\ldots,-3,0,3,6,\ldots\}$ (a subgroup, per the ``Subgroups You Can Picture'' example in Section~3.1). The cosets are:
\begin{itemize}[leftmargin=*]
  \item $0 + 3\mathbb{Z} = \{\ldots,-3,0,3,6,\ldots\}$ (multiples of 3)
  \item $1 + 3\mathbb{Z} = \{\ldots,-2,1,4,7,\ldots\}$ (remainder $1$ when divided by 3)
  \item $2 + 3\mathbb{Z} = \{\ldots,-1,2,5,8,\ldots\}$ (remainder $2$ when divided by 3)
\end{itemize}
Every integer falls into exactly one of these three buckets --- and which bucket it falls into is precisely its remainder mod $3$. There is no fourth distinct coset: $3 + 3\mathbb{Z} = 0 + 3\mathbb{Z}$ (check: $3-0=3 \in 3\mathbb{Z}$), so the pattern repeats.
\end{exbox}

\begin{defbox}{Quotient Group}
The set of all distinct cosets of $H$ in $G$, with addition defined by $(a+H) + (b+H) := (a+b) + H$, is itself a group --- the \textbf{quotient group}, written $G/H$ (``$G$ mod $H$''). Its identity element is the coset $0 + H = H$ itself.
\end{defbox}

\begin{rmkbox}{$\mathbb{Z}_n$ Is Not Just \emph{Like} $\mathbb{Z}/n\mathbb{Z}$ --- It \emph{Is} $\mathbb{Z}/n\mathbb{Z}$}
Compare the coset example above to Section~6: the three cosets of $3\mathbb{Z}$ in $\mathbb{Z}$ correspond \emph{exactly} to the three elements $\{0,1,2\}$ of $\mathbb{Z}_3$, and coset addition matches addition mod $3$ exactly. This is not a coincidence or an analogy --- it is a theorem: $\mathbb{Z}_n$, as defined informally in Section~6, is precisely the quotient group $\mathbb{Z}/n\mathbb{Z}$. ``Wraparound arithmetic'' and ``sorting integers into cosets of $n\mathbb{Z}$'' are two descriptions of the exact same object. This is the rigorous justification for a claim we made without proof in Section~6: that $\mathbb{Z}_n$ really is a well-defined group (in fact ring) under addition and multiplication with wraparound --- it inherits this structure directly from being a quotient of $\mathbb{Z}$.
\end{rmkbox}

\begin{exercisebox}{Practice}
Let $H = 4\mathbb{Z} \subseteq \mathbb{Z}$. List all distinct cosets of $H$ in $\mathbb{Z}$, and confirm there are exactly $4$ of them, matching $\mathbb{Z}_4 = \{0,1,2,3\}$. Which coset does $-5$ belong to?
\end{exercisebox}

\section{Rings}

A ring is what you get when you ask a set to support \emph{two} binary operations that play two structurally different roles --- one behaves like addition (fully invertible), the other behaves like multiplication (weaker, no inverses required). We always \emph{label} these two operations ``$+$'' and ``$\times$,'' but --- and this is worth being precise about, since it's easy to misread --- \textbf{the labels do not mean the operations have to literally be school-arithmetic addition and multiplication}. They can be almost any two operations at all, as long as together they satisfy the axioms below. ``$+$'' and ``$\times$'' are names for the \emph{roles} the two operations play, not a requirement about what the operations must concretely do.

\begin{defbox}{Ring}
A set $R$ with two binary operations $+$ and $\times$ is a \textbf{ring} $(R, +, \times)$ if:
\begin{enumerate}[leftmargin=*, label=(R\arabic*)]
  \item $(R, +)$ is an abelian group (with identity element usually written $0$).
  \item $\times$ is associative: $(a \times b) \times c = a \times (b \times c)$.
  \item The distributive laws hold: $a \times (b + c) = (a \times b) + (a \times c)$ and $(a+b)\times c = (a \times c) + (b \times c)$.
\end{enumerate}
Note: a ring does \emph{not} need multiplicative inverses, and multiplication need not be commutative in general (though in every ring we use in this thesis, it will be --- these are called \textbf{commutative rings}). If there is an element $1 \in R$ with $a \times 1 = 1 \times a = a$ for all $a$, we say $R$ is a \textbf{ring with unity}.
\end{defbox}

\begin{rmkbox}{``$+$'' and ``$\times$'' Are Role Labels, Not a Restriction to Arithmetic}
Compare this to groups (Section~3): a group has \emph{one} operation, and we wrote it as a generic $*$ precisely to emphasize that it could be anything (addition, multiplication, matrix multiplication, function composition --- whatever satisfies the axioms). A ring needs \emph{two} operations, and it turns out we always call them ``$+$'' and ``$\times$'' --- but this is a naming convention for the two \emph{roles}, not a claim that the operations must be ordinary numeric addition and multiplication. Concretely:
\begin{itemize}[leftmargin=*]
  \item \textbf{$n \times n$ matrices:} ``$+$'' is ordinary matrix addition (this forms an abelian group), and ``$\times$'' is matrix \emph{multiplication} (associative, distributes over $+$) --- but matrix multiplication is \textbf{not} commutative in general, which is fine, since a ring's axioms above don't require it (only \emph{commutative} rings require it, and this one wouldn't qualify as one).
  \item \textbf{Boolean rings} (used in digital logic): ``$+$'' is XOR and ``$\times$'' is AND on $\{0,1\}$ --- nothing to do with school arithmetic at all, yet every ring axiom above holds.
\end{itemize}
So whenever you meet a new ring in this course, the first question to ask is not ``is $+$ really addition and is $\times$ really multiplication?'' but rather: \emph{what are the two operations here, and do they satisfy axioms (R1)--(R3)?} The labels ``$+$''/``$\times$'' just tell you which operation is playing which role once you've checked that.
\end{rmkbox}

\begin{exbox}{Rings You Already Know}
\begin{itemize}[leftmargin=*]
  \item $(\mathbb{Z}, +, \times)$ --- the integers form a commutative ring with unity $1$. Notice $2$ still has no multiplicative inverse, and that's fine --- rings don't require one.
  \item \textbf{Polynomial rings}, e.g. $\mathbb{Z}[x]$: the set of all polynomials with integer coefficients (like $3x^2 - x + 7$), with ordinary polynomial addition and multiplication. This is a ring for exactly the same reason $\mathbb{Z}$ is.
\end{itemize}
\end{exbox}

\begin{previewbox}{The Ring That Kyber and Dilithium Actually Use}
Both Kyber and Dilithium do their arithmetic inside a specific ring of the form
$$R_q = \mathbb{Z}_q[x] / (x^n + 1),$$
read as: ``polynomials with coefficients taken modulo $q$ (Section~6), where we also reduce powers of $x$ using the rule $x^n \equiv -1$.'' This looks intimidating right now, but it is built from exactly the pieces in this lecture: a polynomial ring (this section), with coefficients from $\mathbb{Z}_q$ (Section~6), further ``folded'' by a fixed rule. We will unpack this fully in a later lecture --- for now, just notice that ``ring'' is not an abstract curiosity; it's the literal object the cryptography is built on.
\end{previewbox}

\subsection{Zero Divisors: A Structural Reason Some Rings Aren't Fields}

We will soon see (Section~5) that $\mathbb{Z}_n$ is a field exactly when $n$ is prime. Section~6.1 will justify this computationally, via $\gcd$. Here is the deeper, structural reason \emph{why} --- worth seeing now, while we're still inside general ring theory.

\begin{defbox}{Zero Divisor}
In a ring $R$, a nonzero element $a$ is a \textbf{zero divisor} if there exists some \emph{nonzero} $b \in R$ with $a \times b = 0$. A commutative ring with unity that has \emph{no} zero divisors (other than needing $a$ or $b$ to be $0$) is called an \textbf{integral domain}.
\end{defbox}

\begin{exbox}{Zero Divisors in $\mathbb{Z}_6$, Absent in $\mathbb{Z}_5$}
In $\mathbb{Z}_6$: $2 \times 3 = 6 \equiv 0 \pmod 6$, but neither $2$ nor $3$ is $0$. So $2$ and $3$ are both zero divisors --- $\mathbb{Z}_6$ is \emph{not} an integral domain. Contrast this with $\mathbb{Z}_5$: checking every pair of nonzero elements ($1,2,3,4$), no product is ever $\equiv 0 \pmod 5$ (e.g., $2\times 3=6\equiv 1$, $4\times 4 = 16 \equiv 1$, etc.) --- $\mathbb{Z}_5$ has no zero divisors.
\end{exbox}

\begin{rmkbox}{Why This Explains the Prime/Composite Split}
\textbf{Fact:} every field is automatically an integral domain (if $a \ne 0$ and $ab=0$, multiply both sides by $a^{-1}$ to get $b=0$ --- so no nonzero $b$ can pair with a nonzero $a$ to give $0$). Contrapositive: \emph{if a ring has a zero divisor, it cannot be a field.} Now, $\mathbb{Z}_n$ has a zero divisor exactly when $n$ is composite: write $n = jk$ with $1 < j,k < n$; then $j \times k = n \equiv 0 \pmod n$, with neither $j$ nor $k$ equal to $0$ in $\mathbb{Z}_n$. So every composite modulus produces zero divisors, which immediately rules out being a field --- a purely structural explanation, with no $\gcd$ computation needed, of the same fact Section~6.1 reaches computationally.
\end{rmkbox}

\begin{exercisebox}{Practice}
Find a pair of zero divisors in $\mathbb{Z}_{12}$ (there are several valid answers). Then explain, using the Fact above, why $\mathbb{Z}_{12}$ cannot be a field --- without computing any inverses directly.
\end{exercisebox}

\section{Fields}

\begin{defbox}{Field}
A \textbf{field} is a set $F$ with two binary operations $+$ and $\times$ such that:
\begin{enumerate}[leftmargin=*, label=(F\arabic*)]
  \item $(F, +)$ is an abelian group, over \emph{all} of $F$ (identity element $0$).
  \item $(F \setminus \{0\}, \times)$ is an abelian group, over $F$ \textbf{with $0$ removed} (identity element $1$).
  \item The distributive law holds: $a \times (b+c) = (a\times b) + (a \times c)$.
\end{enumerate}
Informally: a field is a set where you can add, subtract, multiply, and divide (by anything except $0$) and always stay inside the set. In short --- \textbf{a field is two abelian groups sharing one set, glued together by distributivity.}
\end{defbox}

\begin{rmkbox}{Why $0$ Has to Be Excluded, and It's Not Optional}
In \emph{any} structure with a distributive law, $0 \times x = 0$ for every $x$ (this follows from distributivity itself, not from an extra rule: $0 \times x = (0+0)\times x = 0\times x + 0 \times x$, so subtracting $0\times x$ from both sides gives $0 = 0 \times x$). So $0$ could only have a multiplicative inverse if $0 \times x = 1$ for some $x$ --- but the left side is always $0$, never $1$ (as long as $1 \ne 0$, which we always assume). So $0$ is permanently excluded from having an inverse, in every field, without exception; it isn't a special case someone chose, it's forced by the arithmetic.
\end{rmkbox}

\begin{defbox}{Subtraction and Division Are \emph{Derived}, Not Extra}
Neither subtraction nor division is a new, independent operation requiring its own axiom --- both are just notation for ``combine with an inverse,'' one level for each operation:
$$
a - b := a + (-b), \qquad \frac{a}{b} := a \times b^{-1} \ \ (b \ne 0),
$$
where $-b$ is guaranteed to exist by $(F,+)$ being a group (F1), and $b^{-1}$ is guaranteed to exist, for every \emph{nonzero} $b$, by $(F\setminus\{0\},\times)$ being a group (F2). So ``a field supports division'' is not an extra requirement on top of the definition --- it falls straight out of (F2) once every nonzero element is already promised an inverse. This is exactly why (F2) excludes $0$: since $0$ can never have an inverse (see the remark above), ``dividing by $0$'' can never be defined this way, no matter what field you're in.
\end{defbox}

\begin{exbox}{Division in $\mathbb{Z}_5$, Worked Out}
What is $4 \div 3$ in $\mathbb{Z}_5$? First find $3^{-1} \bmod 5$: $3 \times 2 = 6 \equiv 1 \pmod 5$, so $3^{-1} = 2$. Then $4 \div 3 := 4 \times 3^{-1} = 4 \times 2 = 8 \equiv 3 \pmod 5$. Check: $3 \times 3 = 9 \equiv 4 \pmod 5$ --- correct, exactly as ``$4 \div 3 = 3$'' should verify.
\end{exbox}

\begin{rmkbox}{The One-Line Takeaway}
\textbf{Subtraction is the additive inverse; division is the multiplicative inverse.} Same trick, applied once per operation --- $a-b$ means ``add $a$ and $b$'s additive inverse,'' $a/b$ means ``multiply $a$ by $b$'s multiplicative inverse.'' The one asymmetry between them traces directly back to (F1) vs.\ (F2): \emph{every} element of $F$ has an additive inverse, since $(F,+)$ is a group over \emph{all} of $F$ --- so subtraction always works. But \emph{only nonzero} elements have a multiplicative inverse, since $(F\setminus\{0\},\times)$ is a group only after removing $0$ --- so division by $b$ only ever works when $b \ne 0$.
\end{rmkbox}

\begin{factbox}{Every Field Is a Ring}
Every field $F$ (Definition above) is automatically a commutative ring with unity --- in fact, an \textbf{integral domain} (Section~4.1). \emph{Why:} (F1) is exactly ring axiom (R1). Associativity and distributivity of $\times$ (ring axioms (R2)--(R3)) are inherited directly from $\times$ being a group operation on $F \setminus \{0\}$ together with (F3). So a field satisfies every ring axiom, plus two extra guarantees stacked on top: $\times$ is commutative, and every \emph{nonzero} element is invertible. This is why some textbooks instead \emph{define} a field as ``a commutative ring with unity in which every nonzero element has a multiplicative inverse'' --- that phrasing and the definition above pick out exactly the same objects; it's just built in the reverse order (starting from ``ring'' and adding a condition, instead of starting from two groups and gluing them together). Finally, a field is automatically an integral domain too: if $ab=0$ with $a \ne 0$, multiplying both sides by $a^{-1}$ forces $b = a^{-1}(ab) = a^{-1}\cdot 0 = 0$, so there can be no zero divisors.
\end{factbox}

\begin{exbox}{Fields and Non-Fields}
\begin{itemize}[leftmargin=*]
  \item $\mathbb{Q}$ (rationals), $\mathbb{R}$ (reals), $\mathbb{C}$ (complex numbers) are all fields --- you can divide by any nonzero element and stay inside the set.
  \item $\mathbb{Z}$ is \textbf{not} a field: $2 \in \mathbb{Z}$ has no multiplicative inverse in $\mathbb{Z}$ (again, $1/2 \notin \mathbb{Z}$). $\mathbb{Z}$ is a ring, but not a field.
  \item $\mathbb{Z}_5 = \{0,1,2,3,4\}$ under addition and multiplication mod $5$ \emph{is} a field --- every nonzero element has an inverse (you checked this, or will check this, in the Section~3 subgroup/group exercise). Fields built this way, from $\mathbb{Z}_p$ with $p$ prime, are called \textbf{finite fields} or \textbf{Galois fields}, written $\mathbb{F}_p$ or $GF(p)$.
\end{itemize}
\end{exbox}

\begin{rmkbox}{The One-Sentence Summary So Far}
Group $\subset$ Ring $\subset$ Field, in the sense that a field is a ring with extra structure (division), and a ring is built from a group with extra structure (a second operation). Each step adds a capability: groups let you undo one operation (inverses); rings add a second, distributive operation; fields guarantee \emph{both} operations are fully invertible (except division by $0$).
\end{rmkbox}

\section{The Modulus: Arithmetic That Wraps Around}

\begin{defbox}{Congruence Modulo $n$}
For a fixed positive integer $n$ (the \textbf{modulus}), two integers $a$ and $b$ are \textbf{congruent modulo $n$}, written
$$a \equiv b \pmod{n},$$
if $n$ divides $(a-b)$ evenly --- equivalently, $a$ and $b$ leave the same remainder when divided by $n$. The set of possible remainders, $\{0, 1, 2, \ldots, n-1\}$, together with addition and multiplication performed ``with wraparound,'' is written $\mathbb{Z}_n$ (or $\mathbb{Z}/n\mathbb{Z}$).
\end{defbox}

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1.1]
  \foreach \i in {0,...,11} {
    \node[circle, draw, minimum size=0.75cm, font=\small] at ({90-\i*30}:2) {\i};
  }
  \draw[-{Latex[length=2mm]}, thick, blue] (2.6,0.3) arc (7:83:2.7);
  \node[blue, font=\small] at (3.2,1.9) {$9 + 5 = 14 \equiv 2 \pmod{12}$};
\end{tikzpicture}
\caption{The clock analogy for modular arithmetic: $\mathbb{Z}_{12}$ works exactly like a 12-hour clock face. $9$ o'clock plus $5$ hours wraps around to $2$ o'clock, i.e., $9+5=14 \equiv 2 \pmod{12}$.}
\label{fig:clock}
\end{figure}

\begin{exbox}{Computing Modulo}
$17 \bmod 5$: divide $17$ by $5$ to get $3$ remainder $2$, so $17 \equiv 2 \pmod 5$. Similarly, $23 \equiv 3 \pmod{10}$, and $-1 \equiv n-1 \pmod n$ (e.g., $-1 \equiv 4 \pmod 5$) --- negative numbers wrap around too.
\end{exbox}

\begin{rmkbox}{When Is $\mathbb{Z}_n$ a Ring? When Is It a Field?}
$\mathbb{Z}_n$ is \emph{always} a commutative ring with unity, for any $n \geq 2$. But it is a \emph{field} only when $n = p$ is \textbf{prime}. If $n$ is composite (e.g., $n=6$), some nonzero elements fail to have multiplicative inverses --- e.g., in $\mathbb{Z}_6$, there is no $x$ with $2x \equiv 1 \pmod 6$, since $2$ shares a common factor with $6$. This is exactly why cryptographic schemes that need a field (rather than just a ring) are built with a prime modulus $q$. (Section~4.1 gave the deeper structural reason: composite $n$ always produces zero divisors, and a ring with zero divisors can never be a field.)
\end{rmkbox}

\subsection{Finding Inverses Without Guessing: The Extended Euclidean Algorithm}

So far, the only way we've found a multiplicative inverse mod $n$ is trial and error (Exercise: ``find $x$ with $3x \equiv 1 \pmod 7$'' --- you could just try $x=1,2,3,\ldots$ until one works). That's fine for small numbers, but real cryptographic moduli are enormous (Kyber's modulus is $q=3329$; RSA moduli have hundreds of digits) --- trial and error is completely impractical. We need an actual algorithm.

\begin{defbox}{Greatest Common Divisor (GCD)}
For integers $a, b$ (not both zero), $\gcd(a,b)$ is the largest positive integer that divides both $a$ and $b$. Two integers are \textbf{coprime} (or relatively prime) if $\gcd(a,b) = 1$.
\end{defbox}

\begin{rmkbox}{Fact: When Does an Inverse Exist?}
An element $a \in \mathbb{Z}_n$ has a multiplicative inverse if and only if $\gcd(a,n) = 1$. This is why every nonzero element of $\mathbb{Z}_p$ (for prime $p$) has an inverse: any $a$ with $1 \le a \le p-1$ automatically satisfies $\gcd(a,p)=1$, since $p$'s only divisors are $1$ and $p$ itself.
\end{rmkbox}

The classical \textbf{Euclidean Algorithm} computes $\gcd(a,b)$ by repeated division with remainder: $\gcd(a,b) = \gcd(b, a \bmod b)$, stopping when the remainder hits $0$. The \textbf{Extended} Euclidean Algorithm does this same process but also tracks, at \emph{every single step}, how the current remainder can be written as a combination $a\cdot x + b\cdot y$ (integers $x,y$) --- and when the algorithm finishes with $\gcd(a,b)=1$, the row where the remainder equals $1$ hands you the inverse directly. The cleanest way to track this --- much easier to follow than plain back-substitution --- is a running \textbf{table} with one row per step.

\begin{defbox}{The Extended Euclidean Algorithm, as a Table}
To find $\gcd(a,b)$ together with integers $x,y$ satisfying $a\cdot x + b\cdot y = \gcd(a,b)$, build a table with columns $q$ (quotient), $r$ (remainder), $x$, $y$, starting with two fixed initial rows and then one new row per division step:
\begin{center}
\begin{tabular}{c|c|c|c|c}
step & $q$ & $r$ & $x$ & $y$ \\
\hline
$-1$ & --- & $a$ & $1$ & $0$ \\
$0$ & --- & $b$ & $0$ & $1$ \\
$1$ & $q_1 = \lfloor r_{-1}/r_0 \rfloor$ & $r_1 = r_{-1} - q_1 r_0$ & $x_1 = x_{-1} - q_1 x_0$ & $y_1 = y_{-1} - q_1 y_0$ \\
$2$ & $q_2 = \lfloor r_0/r_1 \rfloor$ & $r_2 = r_0 - q_2 r_1$ & $x_2 = x_0 - q_2 x_1$ & $y_2 = y_0 - q_2 y_1$ \\
$\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$
\end{tabular}
\end{center}
Every single row satisfies the \textbf{invariant} $r_i = a \cdot x_i + b \cdot y_i$ by construction --- this is the whole mechanism, and it's worth checking explicitly on at least one row the first time you use this table, to convince yourself it isn't magic. Stop once a row produces $r_i = 0$; the \emph{previous} row's $r$ is $\gcd(a,b)$, and that row's $x,y$ are exactly the combination you want.
\end{defbox}

\begin{exbox}{Table Method, Short Version: $\gcd(7,3)$ and $3^{-1} \bmod 7$}
\begin{center}
\begin{tabular}{c|c|c|c|c|l}
step & $q$ & $r$ & $x$ & $y$ & check: $7x+3y$ \\
\hline
$-1$ & --- & $7$ & $1$ & $0$ & $7(1)+3(0)=7$ \checkmark \\
$0$ & --- & $3$ & $0$ & $1$ & $7(0)+3(1)=3$ \checkmark \\
$1$ & $\lfloor 7/3\rfloor=2$ & $7-2(3)=1$ & $1-2(0)=1$ & $0-2(1)=-2$ & $7(1)+3(-2)=1$ \checkmark \\
$2$ & $\lfloor 3/1\rfloor=3$ & $3-3(1)=0$ & $0-3(1)=-3$ & $1-3(-2)=7$ & (stop: $r=0$) \\
\end{tabular}
\end{center}
The row \emph{before} $r$ hit $0$ is step $1$: $r_1 = 1 = \gcd(7,3)$, with $x_1=1, y_1=-2$. So $7(1) + 3(-2) = 1$, i.e., $3 \times (-2) \equiv 1 \pmod 7$, giving $3^{-1} \equiv -2 \equiv 5 \pmod 7$ --- matching what trial and error would also find, but now via a method with a clear, mechanical stopping point.
\end{exbox}

\begin{exbox}{Table Method, Longer Version: $15^{-1} \bmod 26$}
Here the table earns its keep --- more steps than the tiny example above, but the process is identical, row by row.
\begin{center}
\begin{tabular}{c|c|c|c|c}
step & $q$ & $r$ & $x$ & $y$ \\
\hline
$-1$ & --- & $26$ & $1$ & $0$ \\
$0$ & --- & $15$ & $0$ & $1$ \\
$1$ & $\lfloor 26/15\rfloor=1$ & $26-1(15)=11$ & $1-1(0)=1$ & $0-1(1)=-1$ \\
$2$ & $\lfloor 15/11\rfloor=1$ & $15-1(11)=4$ & $0-1(1)=-1$ & $1-1(-1)=2$ \\
$3$ & $\lfloor 11/4\rfloor=2$ & $11-2(4)=3$ & $1-2(-1)=3$ & $-1-2(2)=-5$ \\
$4$ & $\lfloor 4/3\rfloor=1$ & $4-1(3)=1$ & $-1-1(3)=-4$ & $2-1(-5)=7$ \\
$5$ & $\lfloor 3/1\rfloor=3$ & $3-3(1)=0$ & $3-3(-4)=15$ & $-5-3(7)=-26$ \\
\end{tabular}
\end{center}
The row before $r$ hit $0$ is step $4$: $r_4 = 1 = \gcd(26,15)$, with $x_4=-4, y_4=7$. \textbf{Check the invariant directly:} $26(-4) + 15(7) = -104 + 105 = 1$ \checkmark. Reading this modulo $26$: $15 \times 7 \equiv 1 \pmod{26}$, so $15^{-1} \equiv 7 \pmod{26}$. \textbf{Final check:} $15 \times 7 = 105 = 4\times26+1 \equiv 1 \pmod{26}$. Correct.
\end{exbox}

\begin{exercisebox}{Practice}
Use the table method above (not back-substitution, and not trial and error) to find $5^{-1} \bmod 11$. Build the full table --- steps $-1, 0, 1, 2, \ldots$ --- and stop at the correct row. Then check your answer by direct multiplication.
\end{exercisebox}

\begin{previewbox}{Why Cryptography Loves the Modulus}
Modular arithmetic gives us two things cryptography needs badly: (1) a \textbf{finite} number system (so keys and computations have a fixed, bounded size, no matter how much you compute), and (2) built-in \textbf{one-wayness} in places --- e.g., given only $a \bmod n$, you cannot tell which of infinitely many integers $a$ actually was. The Learning-With-Errors (LWE) problem from the proposal document hides its secret inside equations that live in exactly this kind of finite, modular number system (there, the modulus is usually called $q$).
\end{previewbox}

\begin{exercisebox}{Practice}
(a) Compute $23 \times 15 \bmod 7$. (b) List all elements of $\mathbb{Z}_8$ that \emph{do not} have a multiplicative inverse, and explain why (hint: compare each element to $8$). (c) Confirm $\mathbb{Z}_7$ is a field by checking that $3$ has a multiplicative inverse mod $7$ (find $x$ such that $3x \equiv 1 \pmod 7$).
\end{exercisebox}

\section{Vector Spaces, Basis, and Dimension}

\begin{defbox}{Vector Space}
A \textbf{vector space} over a field $F$ is a set $V$ (whose elements are called \textbf{vectors}) equipped with vector addition and scalar multiplication (multiplying a vector by an element of $F$), satisfying the ``obvious'' rules you'd expect: addition makes $(V,+)$ an abelian group, and scalar multiplication distributes sensibly over both vector addition and field addition. The most important example for us: $\mathbb{R}^n$, the set of all $n$-tuples of real numbers, is a vector space over $\mathbb{R}$.
\end{defbox}

\begin{defbox}{Linear Combination, Span, Linear Independence}
Given vectors $v_1, \ldots, v_k \in V$ and scalars $c_1, \ldots, c_k \in F$, the vector $c_1 v_1 + c_2 v_2 + \cdots + c_k v_k$ is called a \textbf{linear combination} of $v_1, \ldots, v_k$. The set of \emph{all} linear combinations of a set of vectors is called their \textbf{span}. The vectors $v_1, \ldots, v_k$ are \textbf{linearly independent} if the only way to write $0 = c_1 v_1 + \cdots + c_k v_k$ is with every $c_i = 0$ --- informally, no vector in the set is ``redundant'' (expressible using the others).
\end{defbox}

\begin{defbox}{Basis and Dimension}
A set of vectors $B = \{b_1, \ldots, b_k\}$ is a \textbf{basis} of a vector space $V$ if (1) $B$ is linearly independent, and (2) $B$ \textbf{spans} $V$ (every vector in $V$ can be written as a linear combination of $B$). Every basis of a given vector space has the same number of elements $k$, called the \textbf{dimension} of $V$.
\end{defbox}

\begin{exbox}{Basis of $\mathbb{R}^2$}
The vectors $e_1 = (1,0)$ and $e_2 = (0,1)$ form the ``standard'' basis of $\mathbb{R}^2$: any vector $(a,b)$ is exactly $a \cdot e_1 + b \cdot e_2$. But this is far from the \emph{only} basis --- $\{(1,1), (1,-1)\}$ is also a valid basis of $\mathbb{R}^2$ (check: they're linearly independent, and any $(a,b)$ can be written as a combination of them). A vector space almost always has infinitely many possible bases; some are more convenient to work with than others. \textbf{Keep this fact in mind --- it becomes very important in Section~8.}
\end{exbox}

\subsection{Inner Products, Norms, and Orthogonality}

To make ``good basis'' (short, nearly-perpendicular vectors) versus ``bad basis'' (long, nearly-parallel vectors) precise rather than just intuitive, we need a way to measure length and angle.

\begin{defbox}{Inner Product and Norm}
For $x = (x_1,\ldots,x_n), y=(y_1,\ldots,y_n) \in \mathbb{R}^n$, the (standard) \textbf{inner product} is $\langle x, y \rangle = \sum_{i=1}^n x_i y_i$. The \textbf{norm} (length) of $x$ is $\|x\| = \sqrt{\langle x,x \rangle}$. Two vectors are \textbf{orthogonal} if $\langle x,y \rangle = 0$ (the angle between them is $90^\circ$). A basis $\{b_1,\ldots,b_k\}$ is an \textbf{orthogonal basis} if every pair $b_i, b_j$ ($i \ne j$) is orthogonal.
\end{defbox}

\begin{exbox}{Computing Norms and Checking Orthogonality}
For $x=(3,4)$: $\|x\| = \sqrt{3^2+4^2} = \sqrt{25} = 5$. For $x=(1,1)$ and $y=(1,-1)$: $\langle x,y\rangle = (1)(1)+(1)(-1) = 0$, so these two vectors (the alternate basis of $\mathbb{R}^2$ from the previous example) are orthogonal --- a nicer basis, geometrically, than it might have looked at first glance.
\end{exbox}

\begin{previewbox}{Quantifying ``Good'' vs.\ ``Bad'' Bases}
With norms in hand, we can finally give a real, measurable meaning to the ``good basis / bad basis'' idea flagged repeatedly in this lecture: a good basis has vectors with small norm $\|b_i\|$ that are close to orthogonal; a bad basis for the \emph{same} lattice can have arbitrarily large norms and near-parallel vectors, despite generating an identical set of points. One common numerical measure of ``how good'' a basis is is its \textbf{orthogonality defect}, $\frac{\prod_i \|b_i\|}{\det(L)}$ (see Section~8.4 below for $\det(L)$) --- it equals $1$ for a perfectly orthogonal basis and grows the more skewed the basis is.
\end{previewbox}

\subsection{Gram--Schmidt Orthogonalization}

Given \emph{any} basis, can we systematically produce an orthogonal one (spanning the same space, though generally \emph{not} the same lattice --- more on that distinction below)? Yes:

\begin{defbox}{Gram--Schmidt Process}
Given linearly independent $b_1,\ldots,b_k$, define $b_1^* = b_1$, and for $i = 2,\ldots,k$:
$$
b_i^* = b_i - \sum_{j=1}^{i-1} \mu_{i,j}\, b_j^*, \qquad \text{where } \mu_{i,j} = \frac{\langle b_i, b_j^* \rangle}{\langle b_j^*, b_j^* \rangle}.
$$
Each $b_i^*$ is $b_i$ with the ``shadow'' it casts on all previous $b_j^*$ subtracted off, leaving only the genuinely new, orthogonal direction. The result $\{b_1^*,\ldots,b_k^*\}$ is an orthogonal basis of the same vector space (though, importantly, its integer combinations generally do \emph{not} produce the same lattice as the original $b_i$'s --- Gram--Schmidt vectors are a vector-space tool, not automatically a lattice basis).
\end{defbox}

\begin{exbox}{Gram--Schmidt on a $2 \times 2$ Example}
Let $b_1 = (2,0)$, $b_2 = (1,2)$. Then $b_1^* = b_1 = (2,0)$. Compute $\mu_{2,1} = \frac{\langle b_2,b_1^*\rangle}{\langle b_1^*,b_1^*\rangle} = \frac{(1)(2)+(2)(0)}{4} = \frac{2}{4} = 0.5$. So $b_2^* = b_2 - 0.5\, b_1^* = (1,2) - (1,0) = (0,2)$. Check: $\langle b_1^*, b_2^* \rangle = (2)(0)+(0)(2) = 0$ --- orthogonal, as required.
\end{exbox}

\begin{previewbox}{Where Gram--Schmidt Leads}
This process reappears, in a modified ``integer-preserving'' form, inside the \textbf{LLL algorithm} --- the classic algorithm for turning a bad lattice basis into a reasonably good one, and a core tool used both to \emph{attack} weak lattice-based schemes and to reason about the security of well-parametrized ones. We'll meet it properly in a later lecture; for now, just remember that Gram--Schmidt is the geometric heart of it.
\end{previewbox}

\begin{exercisebox}{Practice}
Run Gram--Schmidt on $b_1 = (1,1)$, $b_2 = (0,2)$: compute $b_1^*$, $\mu_{2,1}$, and $b_2^*$, and verify $\langle b_1^*, b_2^* \rangle = 0$.
\end{exercisebox}

\section{Lattices, Properly Defined}

Everything so far has allowed \emph{any} scalar from the field $F$ (e.g., any real number) when forming linear combinations. Informally, a lattice is what happens when we impose one dramatic restriction: \textbf{only integer combinations are allowed}, which turns a solid, continuous vector space into a discrete grid of isolated points. That informal picture (Figure~\ref{fig:latticegrid} below) is correct and worth keeping in your head. But the definition we will actually use --- the one used in serious treatments of the subject, such as Chris Peikert's graduate lecture notes \emph{Lattices in Cryptography}\footnote{C. Peikert, \emph{Lattices in Cryptography}, lecture notes, \url{https://github.com/cpeikert/LatticesInCryptography}.} --- starts from the \textbf{subgroup} idea from Section~3.1, not from a basis. We will show the two views agree.

\subsection{Discrete Sets}

\begin{defbox}{Discrete Subset}
A subset $S \subseteq \mathbb{R}^n$ is \textbf{discrete} if every point of $S$ is \emph{isolated}: for every $x \in S$, there exists some $\varepsilon > 0$ such that the open ball $B(x,\varepsilon)$ (all points within distance $\varepsilon$ of $x$) contains \emph{no} point of $S$ other than $x$ itself. Informally: you can always draw a small enough circle around any point of $S$ that catches nothing else from $S$.
\end{defbox}

\begin{rmkbox}{A Shortcut for Subgroups}
Checking every single point of a set for isolation sounds like a lot of work. But if $H$ is a \emph{subgroup} of $(\mathbb{R}^n, +)$ (Section~3.1), there's a shortcut: $H$ is discrete if and only if the origin $\vec{0}$ alone is isolated, i.e., there is some $\varepsilon>0$ with $B(\vec{0},\varepsilon) \cap H = \{\vec{0}\}$. \emph{Why this works:} suppose $\vec 0$ is isolated with radius $\varepsilon$, and suppose some other point $y \in H$ were within $\varepsilon$ of some $x \in H$. Since $H$ is a subgroup, $y - x \in H$ as well, and $y - x$ is within $\varepsilon$ of $\vec 0$ --- so by our assumption, $y - x = \vec 0$, i.e., $y = x$. So checking one point (the origin) is enough for the whole subgroup, thanks to closure under subtraction.
\end{rmkbox}

\subsection{The Definition}

\begin{defbox}{Lattice}
A \textbf{lattice} is a discrete additive subgroup $L$ of $\mathbb{R}^n$. That is, $L \subseteq \mathbb{R}^n$ is a lattice if:
\begin{enumerate}[leftmargin=*, label=(L\arabic*)]
  \item $L$ is a subgroup of $(\mathbb{R}^n, +)$ (Section~3.1): it contains $\vec 0$, and is closed under addition and negation.
  \item $L$ is discrete: by the shortcut above, there exists $\varepsilon > 0$ such that $B(\vec 0, \varepsilon) \cap L = \{\vec 0\}$.
\end{enumerate}
\end{defbox}

\begin{exbox}{$\mathbb{Q}$ Is a Subgroup, But \emph{Not} a Lattice}
Consider $\mathbb{Q} \subseteq \mathbb{R}$ (the rational numbers) under addition. This \emph{is} a subgroup: it contains $0$, the sum of two rationals is rational, and $-a$ is rational whenever $a$ is. So condition (L1) holds. But condition (L2) fails badly: the rationals are \textbf{dense} in $\mathbb{R}$, meaning every open ball around $0$, no matter how tiny, contains infinitely many other rational numbers. There is no $\varepsilon>0$ that isolates $0$. So $\mathbb{Q}$ is a perfectly good subgroup that is \textbf{not} a lattice --- this is exactly why condition (L2) has to be part of the definition; subgroup alone is not a strong enough requirement.
\end{exbox}

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1]
  % Discrete side
  \draw[thick] (-4.2,-0.4) -- (0.2,-0.4);
  \node at (-2,-1.0) {\small $\mathbb{Z}$ --- discrete: points have gaps};
  \foreach \x in {-4,-3,-2,-1,0} {
    \fill[blue] (\x,-0.4) circle (2.2pt);
  }
  \draw[-{Latex[length=2mm]}, teal, thick] (-2.35,-0.05) arc (200:340:0.35);
  \node[teal, font=\tiny] at (-2,0.35) {no other point this close};
  % Dense side
  \draw[thick] (0.8,-0.4) -- (5.2,-0.4);
  \node at (3,-1.0) {\small $\mathbb{Q}$ --- dense: no gaps, ever};
  \foreach \x in {1,1.3,1.5,1.65,1.75,1.82,1.87,1.91,1.94,1.96,1.98,2,2.3,2.6,2.9,3.2,3.5,3.8,4.1,4.4,4.7,5} {
    \fill[red] (\x,-0.4) circle (1.6pt);
  }
  \draw[dashed, teal, thick] (1.7,-0.1) circle (0.35);
  \node[teal, font=\tiny] at (1.7,0.35) {always more points inside};
\end{tikzpicture}
\caption{Discreteness, visualized. Left: in $\mathbb{Z}$, every point has a gap of size $1$ around it with nothing else nearby --- discrete. Right: in $\mathbb{Q}$, no matter how small a circle you draw around a point, there are always infinitely many more rational points crammed inside it --- dense, and therefore \emph{not} discrete.}
\label{fig:discrete}
\end{figure}

\begin{exbox}{$\mathbb{Z}$ and $\mathbb{Z}^n$ Are Lattices}
$\mathbb{Z} \subseteq \mathbb{R}$ is a subgroup (trivially --- and recall from Section~3.1 that $n\mathbb{Z}$ is a subgroup of $\mathbb{Z}$ for any $n$; here we're just looking at $\mathbb{Z}$ itself sitting inside $\mathbb{R}$), and it is discrete: taking $\varepsilon = 0.5$, $B(0, 0.5) \cap \mathbb{Z} = \{0\}$, since the nearest other integers ($\pm 1$) are a full unit away. So $\mathbb{Z}$ is a lattice in $\mathbb{R}$ --- the simplest possible example. The same argument, coordinate by coordinate, shows $\mathbb{Z}^n \subseteq \mathbb{R}^n$ (all integer-coordinate points) is a lattice in $\mathbb{R}^n$; it is often called \textbf{the integer lattice}.
\end{exbox}

\subsection{Recovering the Basis Picture}

The subgroup definition is more general and more standard, but the ``grid generated by a basis'' picture you may already have in your head is not wrong --- it is a theorem, not a definition, and it's worth seeing why.

\begin{factbox}{Basis Representation of Lattices}
Let $L \subseteq \mathbb{R}^n$. Then $L$ is a lattice (in the sense of the subgroup definition above, i.e., a discrete additive subgroup) \textbf{if and only if} there exist linearly independent vectors $b_1, \ldots, b_k \in \mathbb{R}^n$ (with $k \le n$) such that
$$
L = \mathcal{L}(b_1,\ldots,b_k) = \left\{ \sum_{i=1}^{k} a_i b_i \;\middle|\; a_i \in \mathbb{Z} \right\}.
$$
The vectors $b_1,\ldots,b_k$ are called a \textbf{basis} of $L$; the number $k$ is the \textbf{rank} of $L$. If $k=n$ (the basis spans all of $\mathbb{R}^n$), $L$ is called \textbf{full-rank} --- this is the common case in cryptography.
\end{factbox}

\begin{rmkbox}{Proof Idea (Sketch Only)}
\textbf{Easy direction ($\Leftarrow$):} given linearly independent $b_1,\ldots,b_k$, the set of integer combinations is clearly a subgroup (adding two integer combinations, or negating one, gives another integer combination). Discreteness holds because the $b_i$ point in genuinely independent directions: you cannot make $\sum a_i b_i$ arbitrarily close to $\vec 0$ without every $a_i$ being exactly $0$ --- linear independence rules out any ``near-cancellation.'' (Making this fully rigorous uses the fact that the linear map $(a_1,\ldots,a_k) \mapsto \sum a_i b_i$ has a bounded inverse on its image.)

\textbf{Harder direction ($\Rightarrow$):} given only that $L$ is a discrete subgroup, we must \emph{construct} a basis. The idea is greedy and inductive: pick $b_1$ to be a shortest nonzero vector in $L$ (this exists precisely \emph{because} $L$ is discrete --- in a non-discrete set like $\mathbb{Q}$, there is no shortest nonzero element, since you can always find a smaller one). Then look at the part of $L$ not lying on the line through $b_1$, and pick $b_2$ to be a shortest vector there, and so on. A careful argument (using discreteness at each step, and projecting onto the space orthogonal to the vectors already chosen) shows this process terminates with a valid basis after $k \le n$ steps. The full details are carried out rigorously in Peikert's notes; we take the result as given here.
\end{rmkbox}

\begin{previewbox}{A Name for the ``Shortest Vector''}
The shortest nonzero vector used in the proof sketch above has a standard name: $\lambda_1(L)$, the \textbf{minimum distance} of the lattice. Computing $\lambda_1(L)$ --- or even just estimating it --- for a lattice given only a (possibly bad) basis is called the \textbf{Shortest Vector Problem (SVP)}, and it is believed to be extremely hard in high dimensions. This hardness is one of the two or three core hard problems that lattice-based cryptography's security ultimately rests on. We'll return to this once we reach the LWE problem.
\end{previewbox}

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.62]
  \foreach \x in {0,...,7}
    \foreach \y in {0,...,5}
      \fill[gray!70] (\x,\y) circle (1.6pt);
  \draw[-{Latex[length=2mm]}, thick, blue] (2,1) -- (4,2);
  \node[blue, font=\small] at (3.3,2.5) {basis vector $b_1$};
  \draw[-{Latex[length=2mm]}, thick, blue] (2,1) -- (1,3);
  \node[blue, font=\small, left] at (1,3) {basis vector $b_2$};
\end{tikzpicture}
\caption{A 2-dimensional, full-rank lattice generated by a basis $\{b_1, b_2\}$: every grid dot is reachable by some \emph{integer} combination $a_1 b_1 + a_2 b_2$ (with $a_1, a_2 \in \mathbb{Z}$), starting from any chosen origin dot. This is the same object as the subgroup definition above --- just viewed through its basis instead of as an abstract subgroup.}
\label{fig:latticegrid}
\end{figure}

\begin{rmkbox}{Same Lattice, Different Basis}
A single lattice has \emph{many} different valid bases (just as a vector space does --- recall the ``Basis of $\mathbb{R}^2$'' example in Section~7). Two bases generate the exact same set of points if and only if one can be converted to the other using an integer matrix with determinant $\pm 1$ (a \textbf{unimodular} transformation) --- we won't need the details of this fact yet, but the consequence is critical: some bases consist of short, nearly-perpendicular vectors (a ``good'' basis, easy to compute with), while other, equally valid bases for the \emph{same} lattice consist of long, nearly-parallel vectors (a ``bad'' basis, hard to compute with). \textbf{This gap between good and bad bases for the same lattice is the single most important idea in lattice-based cryptography}: the secret key is typically a good (short) basis, while the public key describes the same lattice using a bad (long, scrambled-looking) basis. Recovering the good basis from the bad one is believed to be extremely hard once the dimension $n$ is large enough.
\end{rmkbox}

\begin{exbox}{A Tiny Lattice by Hand}
Let $b_1 = (2,0)$ and $b_2 = (0,3)$ in $\mathbb{R}^2$. The lattice $\mathcal{L}(b_1,b_2)$ consists of every point $(2a_1, 3a_2)$ for integers $a_1, a_2$ --- so it includes $(0,0)$, $(2,0)$, $(-2,0)$, $(0,3)$, $(2,3)$, $(4,-3)$, and so on, but \emph{not} $(1,1)$ or $(1,0)$, since those cannot be written as $(2a_1, 3a_2)$ for integer $a_1, a_2$. As a sanity check against the subgroup definition directly: is this set a subgroup? Yes --- closed under addition and negation, contains $(0,0)$. Is it discrete? Yes --- the nearest point to the origin is at distance $2$, so any $\varepsilon < 2$ isolates $\vec 0$.
\end{exbox}

\subsection{The Determinant (Volume) of a Lattice}

One more piece of structure will matter a great deal in later lectures: every lattice has a well-defined ``volume,'' independent of which basis you use to describe it.

\begin{defbox}{Fundamental Domain and Determinant}
Given a basis $b_1,\ldots,b_n$ of a full-rank lattice $L \subseteq \mathbb{R}^n$, the \textbf{fundamental parallelepiped} is
$$
\mathcal{P}(B) = \left\{ \sum_{i=1}^n t_i b_i \;\middle|\; 0 \le t_i < 1 \right\},
$$
i.e., the region ``tiled out'' by taking all \emph{fractional} (not just integer) combinations between $0$ and $1$. Copies of $\mathcal{P}(B)$, shifted by every lattice point, tile all of $\mathbb{R}^n$ with no gaps or overlaps. The \textbf{determinant} of the lattice, $\det(L)$, is the volume of $\mathcal{P}(B)$, computed as $\det(L) = |\det(B)|$, where $B$ is the matrix whose rows (or columns) are $b_1,\ldots,b_n$.
\end{defbox}

\begin{rmkbox}{Why $\det(L)$ Doesn't Depend on Which Basis You Pick}
Recall from the ``Same Lattice, Different Basis'' remark above: any two bases of the same lattice are related by a unimodular matrix $U$ (integer entries, $\det(U) = \pm 1$), i.e., $B' = UB$. Then $\det(B') = \det(U)\det(B) = \pm \det(B)$, so $|\det(B')| = |\det(B)|$. This confirms $\det(L)$ is a genuine property of the lattice itself, not an artifact of which basis happened to be used to compute it --- exactly the kind of basis-independence you should expect from something called ``the volume of the lattice.''
\end{rmkbox}

\begin{exbox}{Computing $\det(L)$ for the Tiny Lattice}
For $b_1=(2,0)$, $b_2=(0,3)$ from the example above, $B = \begin{pmatrix} 2 & 0 \\ 0 & 3 \end{pmatrix}$, so $\det(L) = |\det(B)| = |2 \times 3 - 0 \times 0| = 6$. Geometrically: the fundamental parallelepiped is just the $2\times 3$ rectangle $[0,2)\times[0,3)$, and copies of this rectangle exactly tile the plane, anchored at every lattice point --- matching the picture in Figure~\ref{fig:latticegrid}'s style, just for this specific (axis-aligned) lattice.
\end{exbox}

\begin{previewbox}{Minkowski's Theorem: Connecting Volume Back to $\lambda_1(L)$}
Here is a genuinely deep fact we will only preview, not prove: \textbf{Minkowski's First Theorem} says that any symmetric convex region of volume greater than $2^n \det(L)$ must contain a nonzero lattice point. Applying this to a ball gives a bound relating the two quantities we've now defined: $\lambda_1(L) \le \sqrt{n} \cdot \det(L)^{1/n}$ (up to constants). In plain terms: a lattice cannot ``hide'' an arbitrarily long shortest vector if its volume is small --- volume and minimum distance are locked together by the geometry, no matter which basis you're handed. This is part of \emph{why} SVP is believed hard only in the ``regime'' cryptographers carefully choose (high dimension $n$, appropriately scaled $\det(L)$) rather than for arbitrary lattices --- a preview of parameter selection, which we'll return to once we discuss how Kyber and Dilithium choose their concrete dimensions and moduli.
\end{previewbox}

\begin{exercisebox}{Practice}
For the lattice generated by $b_1 = (1,2)$, $b_2 = (3,1)$ (from the earlier homework exercise): compute $\det(L)$ directly from the basis matrix. Then pick any \emph{other} valid basis for the same lattice (e.g., $b_1' = b_1$, $b_2' = b_2 - b_1$) and confirm you get the same $\det(L)$.
\end{exercisebox}

\begin{previewbox}{Where This Leads Next Lecture}
With Groups $\to$ Subgroups $\to$ Rings $\to$ Fields $\to$ Modular arithmetic $\to$ Vector spaces $\to$ Basis $\to$ Lattice now in place, the next lecture can finally state the \textbf{Learning With Errors (LWE)} problem precisely: a secret vector, hidden inside noisy linear equations over $\mathbb{Z}_q$ (a ring, Section~6), which is exactly the kind of ``bad basis'' hardness described above, dressed up in a specific, standardized cryptographic form. We will also unpack the ring $R_q = \mathbb{Z}_q[x]/(x^n+1)$ flagged in Section~4, since \textbf{module lattices} (lattices whose coordinates come from a ring like $R_q$ instead of plain integers) are exactly what Kyber and Dilithium use in practice.
\end{previewbox}

\section{Summary Table}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Structure} & \textbf{Requires} & \textbf{Example} \\
\hline
Group & 1 operation, inverses exist & $(\mathbb{Z}, +)$ \\
Subgroup & A subset that is itself a group & $2\mathbb{Z} \subseteq \mathbb{Z}$ \\
Coset / Quotient group & $G$ sorted into shifted copies of $H$ & $\mathbb{Z}/3\mathbb{Z} = \{0,1,2\}$ \\
Ring & 2 operations, no division needed & $(\mathbb{Z}, +, \times)$, $\mathbb{Z}[x]$ \\
Integral domain & Ring, no zero divisors & $\mathbb{Z}$, $\mathbb{Z}_5$ (not $\mathbb{Z}_6$) \\
Field & Ring + division by nonzero elements & $\mathbb{Q}, \mathbb{R}, \mathbb{Z}_p$ ($p$ prime) \\
Modulus & Wraparound arithmetic & $\mathbb{Z}_n = \{0,\ldots,n-1\}$ \\
Vector space & Field + vectors + scalar mult. & $\mathbb{R}^n$ \\
Basis & Independent, spanning vector set & $\{(1,0),(0,1)\}$ in $\mathbb{R}^2$ \\
Orthogonal basis & Basis, all pairs perpendicular & Gram--Schmidt output \\
Lattice & Discrete additive subgroup of $\mathbb{R}^n$ & $\mathbb{Z}^n$; grid of points \\
$\det(L)$ & Volume of the fundamental domain & $|\det(B)|$, any basis $B$ \\
\hline
\end{tabular}
\end{center}

\section{Full List of Exercises (Assign as Homework Before Lecture 2)}

\begin{exercisebox}{Homework Set}
\begin{enumerate}[leftmargin=*]
  \item Verify $(\{1,2,3,4\}, \times \bmod 5)$ is a group by writing out its full multiplication table (Section~3 exercise, above).
  \item Is $3\mathbb{Z} = \{\ldots,-6,-3,0,3,6,\ldots\}$ a subgroup of $(\mathbb{Z},+)$? Justify using the subgroup test (Section~3.1). Is $\{0,3,6,9,\ldots\}$ (only the non-negative multiples of $3$) a subgroup of $(\mathbb{Z},+)$? Why or why not?
  \item List all distinct cosets of $4\mathbb{Z}$ in $\mathbb{Z}$, confirm there are exactly $4$, and identify which coset $-5$ belongs to (Section~3.2 exercise, above).
  \item Find a pair of zero divisors in $\mathbb{Z}_{12}$, and use the zero-divisor fact (Section~4.1) to explain why $\mathbb{Z}_{12}$ cannot be a field, without computing any inverses.
  \item Compute $23 \times 15 \bmod 7$; list the non-invertible elements of $\mathbb{Z}_8$; use the Extended Euclidean Algorithm (Section~6.1) to find the inverse of $3 \bmod 7$ --- confirm it matches the worked example.
  \item Use the Extended Euclidean Algorithm to find $5^{-1} \bmod 11$ (Section~6.1 exercise, above), showing all steps.
  \item Is $\mathbb{Z}_9$ a field? Why or why not? Answer it two ways: (a) using $\gcd$ (Section~6), and (b) using zero divisors (Section~4.1).
  \item Show that $\{(1,1), (1,-1)\}$ is a valid basis of $\mathbb{R}^2$ by (a) checking linear independence and (b) writing the vector $(4,2)$ as a linear combination of them.
  \item Run Gram--Schmidt on $b_1=(1,1)$, $b_2=(0,2)$ (Section~7.2 exercise, above), and verify the result is orthogonal.
  \item For the lattice generated by $b_1=(1,2)$ and $b_2=(3,1)$: (a) is the point $(7,5)$ in the lattice? (b) compute $\det(L)$ from this basis; (c) pick a different valid basis for the same lattice and confirm $\det(L)$ is unchanged (Section~8.4 exercise, above).
  \item Let $L = \{(x,y) \in \mathbb{Z}^2 : x + y \text{ is even}\}$. (a) Show $L$ is a subgroup of $(\mathbb{R}^2, +)$. (b) Show $L$ is discrete (find an $\varepsilon$ that isolates $\vec 0$). (c) Since $L$ is therefore a lattice, find a basis $\{b_1, b_2\}$ for it, and compute $\det(L)$.
  \item In your own words (3--4 sentences): why does the definition of a lattice need \emph{both} ``subgroup'' and ``discrete''? Use the $\mathbb{Q}$ example from Section~8.2 to explain what goes wrong if you drop the discreteness requirement.
  \item In your own words (3--4 sentences), explain why a ``good basis'' and a ``bad basis'' for the same lattice can make the same mathematical object feel completely different to work with computationally, now that you can measure this with norms (Section~7.1).
\end{enumerate}
\end{exercisebox}

\end{document}
